### 2021-06-03 nand_lb.tex
writeup

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{amsfonts, amsmath, amsthm, amssymb}
\usepackage[margin=2.5cm]{geometry}
\usepackage{enumitem}

\title{NAND formulas lower bounds}

\newcommand{\NN}{\mathbb{N}}%
\newcommand{\Sbb}{\mathbb{S}}%
\newcommand{\Cc}{\mathcal{C}}%
\newcommand{\Gg}{\mathcal{G}}%
\newcommand{\Ss}{\mathcal{S}}%
\newcommand{\Ot}{\tilde{O}}%
\newcommand{\Poly}{\textsf{P}}%
\newcommand{\NP}{\textsf{NP}}%
\newcommand{\bit}{\{0,1\}}%
\DeclareMathOperator*{\poly}{poly}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
% \newtheorem{problem}{Problem}
\newtheorem{prop}[theorem]{Proposition}
\newtheorem{property}[theorem]{Property}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}{Remark}

\begin{document}

\maketitle

We consider formulas with gates of fan-in 2.

\section{Shallow NAND formula for every boolean function}

\begin{definition}
	The multiplexer function with parameter $n$ is
	the function defined by:
	\[MUX_{n}: \begin{cases}
	\bit^{n+2^n} \rightarrow \bit \\
	(x_1,\ldots,x_n, r_1,\ldots, r_{2^n}) \mapsto r_t, \text{ where } x_1\ldots x_n \text{ is the binary representation of } t
	\end{cases}
	\]
\end{definition}

\begin{lemma}\label{lemma:shallow-mux}
	The $MUX_{n}$ function has a NAND formula of depth $n + 2\log n + O(1)$.
\end{lemma}
\begin{proof}
	If the structure was unrestricted, we could simply build a binary tree of OR gates of depth $n$,
	and put at each leaf a formula of depth $\log(n)$ which is equivalent to $f(y) \wedge x = y$,
	where $x$ denotes the vector $x_1,\ldots,x_n$. In that case, for a given input $x$, each of these formula except at most one is equal to zero, and if there is a one, it ``flows'' up to the output gate via the OR tree.

	The idea is to do something similar, except that the result has to be propagated to the root through a OR/AND tree. We can achieve this by ``activating'' some other well-chosen formulas.
	We use a top-down approach, starting from the output OR gate. There is nothing needed to propagate the result through OR gates. For AND gate, the results comes on one wire, and there must be a 1 on the other wire.

	In mathematical terms: we associate to each node of the OR/AND tree a bit string of length $\leq n$ 
	in the following way: 
	start with the empty string at the root, append a zero when you move to a left child, and a one when you move to a right child. Leaves have bit strings of length $n$.
	Given a leaf with bitstring $y$, we add below it the formula 
	$((y = x) \wedge f(y)) \vee ((y\neq= x) \wedge y fiddle x)$ TODO.
\end{proof}

\begin{corollary}\label{cor:shallow-nand}
	Every boolean function on $n$ variables has NAND formulas of depth $n + 2\log n + O(1)$.
\end{corollary}
\begin{proof}
	We can write $f(x_1,\ldots,x_n)$ as $MUX_{n}(x_1,\ldots, x_n, f(0,\ldots,0),\ldots,f(1,\ldots,1))$.
	By interpreting each $f(b_1,\ldots,b_n)$ as a constant (i.e. has a formula of constant depth), we get the desired result.
\end{proof}

\section{Andreev's function(s) have NAND formulas of size $\Ot(n^3)$}

Andreev's function with parameters $k, m$
is the function $A_{k,m} : \bit^{\bit^k} \times (\bit^m)^k \rightarrow \bit$
which maps a boolean function $f: {\bit^k} \rightarrow \bit$ 
(given as a bit string of length $2^k$ representing its truth table)
and $k$ bit strings $(X_1,\ldots, X_k)$ of length $m$
to $f(X_{11} \oplus \ldots \oplus X_{1m}, \ldots, X_{k1} \oplus \ldots \oplus X_{km})$.
$A_{k,m}$ can be seen as a boolean function which takes $n = 2^k + km$ inputs.
In can be shown that $A_{k,m}$ does not have (classical De Morgan's) formulas
of size $O(n^{3-\epsilon})$, for every $\epsilon > 0$.
This lower bound is tight: one can easily construct De Morgan formulas of $\Ot(n^3)$ size
for $A_{k,m}$.

In what follows, we show that $A_{k,m}$ also has NAND formulas of size $\Ot(n^3)$,
which shows that NAND formulas are rather expressive.

Using a simple inductive argument, one can show that any function on $n$ variables has 
NAND formulas of depth $2n$, but this construction is not sufficient to obtain the aforementioned 
formulas for Andreev's function.
Using Lemma~\ref{lemma:shallow-mux}, we can prove the following result.

\begin{corollary}
	Andreev's functions have NAND formulas of depth $3\log n + O(\log \log n)$
	(which have size $\Ot(n^3)$, since each gate has fan-in 2).
\end{corollary}
\begin{proof}
	First, notice that both $a\oplus b$ and $\neg (a\oplus b)$ have NAND formulas of depth 2.
	Therefore, the $XOR$ of $m$ variables can be computed with tree-like NAND formula of depth $2\log m + O(1)$.
	By plugging these formulas applied to each $X_i$ 
	into a NAND formula of depth $k + 2\log k + O(1)$ for $f$ given by Corollary~\ref{cor:shallow-nand},
	we get a that $A_{k,m}$ has a NAND formula of depth 
	$2\log m + k + 2\log k + O(1) \leq  3\log n + 2\log\log n + O(1)$.
\end{proof}

\section{To keep in mind for later}
It seems that function similar to $x_0 \vee (x_1 \wedge (x_2 \vee \ldots))$
are ``hard'', i.e. they have NAND formulas of depth at least $n$ (or $\Omega(n)$ ?).

\begin{itemize}
	\item Is this true ?
	\item Can we formalize it ?
	\item Can we use it to find another hard function like Andreev's ?
\end{itemize}

\end{document}

### 2021-06-10 nand_lb.tex
Slowdown Lemam

@@
 \DeclareMathOperator*{\poly}{poly}
+\DeclareMathOperator*{\polylog}{polylog}
 
@@
 	start with the empty string at the root, append a zero when you move to a left child, and a one when you move to a right child. Leaves have bit strings of length $n$.
-	Given a leaf with bitstring $y$, we add below it the formula 
+	Given a leaf with bit string $y$, we add below it the formula 
 	$((y = x) \wedge f(y)) \vee ((y\neq= x) \wedge y fiddle x)$ TODO.
@@
 
+\section{Speedup Lemma}
+
+\begin{lemma}\label{speedup-lemma}
+	Let $F$ be a uniform OR/AND formula on $n$ variables of depth $d= c \log n$.
+	There is an (co-)nondeterministic algorithm that decides
+	whether an input $v$ satisfies $F$ in time $n^{c/2}\polylog(n)$.
+\end{lemma}
+\begin{proof}
+	We first prove the nondeterministic case.
+	We can assume w.l.o.g. that $d$ is even, and $F$ has depth $d$ on every path.
+
+	Given an OR/AND formula $F$ of depth $d$ a bit string $x$ of length at most $d$, let $F_{|x}$
+	be the sub-formula defined by the subtree starting at the node reached by following the path specified by $x$ from the root or $F$.
+
+	Consider the following recursive algorithm $A$, taking as input an OR/AND formula $F$ of depth $d > 0$, a bit string $x$, and the valuation $v$:
+	\begin{itemize}
+		\item 
+	if $|x| < d$
+	nondeterministically guess $b\in \bit$,
+	and return $A(F, xb0, v) \wedge A(F, xb1, v)$.
+		\item 
+	if $|x| = d$, run the the algorithm $P_F$ on $x$, and return the truth value of the corresponding literal in $v$.
+	\end{itemize}
+
+	Correctness: we show that there exists a $b\in \bit$ such $A(F, x, v)$ returns true iff and only if
+	the formula $F_{|x}(v)$ is true.
+	This is follows by definition when $|x| = d$.
+	By induction, when $|x| < d$, $A(F, xb0, v) \wedge A(F, xb1, v)$ is the value of the AND/OR sub-formula $F_{|xb}$.
+As $F_{|x}$ is an OR/AND formula, it is satisfiable iff at least one of $F_{|x0}$ or $F_{|x1}$ is sat, i.e.
+iff there exists $b\in \bit$ such that $A(F, xb0, v) \wedge A(F, xb1, v)$ is true.
+	The value of $F$ is then given by $A(F, \varepsilon, v)$, where $\varepsilon$ is the empty word.
+
+	Complexity: $A$ makes two recursive calls with the length of $x$ increased by $2$, 
+	and makes a call to $P_F$ at length $d$.
+	Therefore there are $2^{d/2} = n^{c/2}$ calls to $P_F$, and $A$ runs in time $n^{c/2}\polylog(n)$.
+	We also see here that increasing the depth of $F$ by a constant only adds a constant factor in the runtime, hence we can assume that $F$ has even depth.
+
+	For the co-nondeterministic case, we proceed similarly, except that we guess universally on the AND gates,
+	and check both sides of OR gates.
+\end{proof}
+
+\section{Lower bounds}
+
 \section{To keep in mind for later}
-It seems that function similar to $x_0 \vee (x_1 \wedge (x_2 \vee \ldots))$
+It seems that functions of the form $x_0 \vee (x_1 \wedge (x_2 \vee \ldots))$
 are ``hard'', i.e. they have NAND formulas of depth at least $n$ (or $\Omega(n)$ ?).
@@
 	\item Is this true ?
-	\item Can we formalize it ?
-	\item Can we use it to find another hard function like Andreev's ?
+	\item Can we use it to construct a hard function like Andreev's ?
 \end{itemize}

### 2022-05-16 article/main.tex
Draft outline of the paper

\documentclass[a4paper, 12pt]{article}
% \usepackage{ae,lmodern}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[USenglish]{babel}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{todonotes}
\usepackage{enumitem}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section] % Number within sec ?
% \newtheorem{theorem}{Theorem}
\newtheorem{prop}[theorem]{Proposition}
\newtheorem{property}[theorem]{Property}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{observation}[theorem]{Observation}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
% \newtheorem*{remark*}{Remark}
% \theoremstyle{definition}
% \newtheorem{problem}{Problem}
% \newtheorem{openproblem}{Open Problem}

\newcommand{\NN}{\mathbb{N}}%
\newcommand{\ZZ}{\mathbb{Z}}%
\newcommand{\Sbb}{\mathbb{S}}%
\newcommand{\Cc}{\mathcal{C}}%
\newcommand{\Gg}{\mathcal{G}}%
\newcommand{\Ss}{\mathcal{S}}%
\newcommand{\eps}{\varepsilon}%
\newcommand{\epsm}{\eps^{-1}}%
\newcommand{\bit}{\{0,1\}}%
\DeclareMathOperator*{\poly}{poly}

\newcommand{\Poly}{\textsf{P}}%
\newcommand{\NP}{\textsf{NP}}%
\newcommand{\SAT}{\textsf{SAT}}%
\newcommand{\NAND}{\textsf{NAND}}%

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\title{Alternating formula lower bounds for SAT}
\author{Gabriel Bathie, Ryan Williams}
\begin{document}
\maketitle

\begin{abstract}
\end{abstract}

List of things to do:
\begin{itemize}
	\item Improve Spira/Brent to show that a $c \log n$depth 
	lower bound implies an $n^{c-\eps}$ size lower bound?
	\item Show that this bound is optimal w.r.t. to these rules?
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

Motivation

Related work:
- Original line of work : \cite{fortnow2000time,fortnow2005time}
- Main line of work \cite{williams2006inductive,williams2007time,williams2013alternation} and its dramatic resolution \cite{buss2015limits}.
- Broader classes \cite{mudigonda2020time}

Main results:
\begin{theorem}
	For any $c < 4 \cos(\pi/7)$, \SAT does not have \NAND formulas of depth $(c + o(1)) \log n$.
\end{theorem}

\todo[inline]{Prove all the following}
\begin{lemma}
	For any \NAND formula on $n$ inputs of size $n^c$ and any $\eps > 0$, 
	there exists an equivalent \NAND formula of depth $(c+\eps) \log n$.
\end{lemma}

\begin{corollary}
	For any $c < 4 \cos(\pi/7)$, \SAT does not have \NAND formulas of \textbf{size} $n^{c + o(1)}$.
\end{corollary}

\begin{theorem} (??)
	With our rules, we cannot do better than $4 \cos(\pi/7)$.
\end{theorem}

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\section{Definitions}
Notations?

Uniform \NAND/alternating formulas. Specific model, comparison with the usual model.

Alternating computation.

Approach: proof by contradiction.

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\section{Proof of thm}
Speedup Rule for Alternating formulas.

How to get the contradiction given the slowdown rule for $c < 4\cos(\pi /7)$.

\subsection{Improving Brent/Spira}
\todo[inline]{Prove this}

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\section{Proof of optimality}
\todo[inline]{Prove this}

%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{plain}
\bibliography{biblio.bib}

\end{document}

### 2022-05-18 article/main.tex
Added todo

@@
 
+\newcommand{\Omegat}[1]{\widetilde{\Omega}\left( #1 \right)}%
+\newcommand{\Ot}[1]{\widetilde{O}\left( #1 \right)}%
+
 \newcommand{\Poly}{\textsf{P}}%
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\title{Alternating formula lower bounds for SAT}
+\title{Uniform formula lower bounds for finding and counting satisfiable assignments}
 \author{Gabriel Bathie, Ryan Williams}
@@
 \begin{itemize}
-	\item Improve Spira/Brent to show that a $c \log n$depth 
+	\item Improve Spira/Brent to show that a $c \log n$ depth 
 	lower bound implies an $n^{c-\eps}$ size lower bound?
-	\item Show that this bound is optimal w.r.t. to these rules?
+	\item Show that this bound is optimal w.r.t. these rules?
 \end{itemize}
@@
 
-Motivation
+Motivation: understanding complexity. Explicit lower bounds : programs, circuits, formulas.
+Uniformity.
 
@@
 
-Main results:
+Best known formula lower bounds for languages (functions?) in NP: 
+\begin{itemize}
+	\item $\Omegat{n^3}$ (for both uniform and non-uniform formulas) for Andreev's function \cite{hastad1998shrinkage}\todo{check source} in the De Morgan basis $\{\wedge, \vee, \neg\}$.
+	\item $\Omega{n^2/\log n}$ in the full binary basis (Nechiporuk)
+	\item $n^{\Theta(\log n)}$ for monotone formulas (formulas over the $\{\wedge, \vee\}$ basis).
+\end{itemize}
+
+\subsection{Results overview}
+
+We first study \NAND{} formulas. 
+The \NAND{} basis is rather expressive: Andreev's function has \NAND{} formulas of depth $(3+o(1)) \log n$ (and therefore of size $\Ot{n^3}$), which matches the lower bound for the De Morgan Basis.
+
 \begin{theorem}
-	For any $c < 4 \cos(\pi/7)$, \SAT does not have \NAND formulas of depth $(c + o(1)) \log n$.
+	For any $c < 4 \cos(\pi/7) \simeq 3.603$, \SAT does not have \NAND{} formulas of depth $(c + o(1)) \log n$.
 \end{theorem}
 
-\todo[inline]{Prove all the following}
+\todo[inline]{Prove all of the following.
+
+After checking \cite{sergeev2019relation}, this is probably incorrect...}
 \begin{lemma}
-	For any \NAND formula on $n$ inputs of size $n^c$ and any $\eps > 0$, 
-	there exists an equivalent \NAND formula of depth $(c+\eps) \log n$.
+	For any \NAND{} formula on $n$ inputs of size $n^c$ and any $\eps > 0$, 
+	there exists an equivalent \NAND{} formula of depth $(c+\eps) \log n$.
 \end{lemma}
@@
 \begin{corollary}
-	For any $c < 4 \cos(\pi/7)$, \SAT does not have \NAND formulas of \textbf{size} $n^{c + o(1)}$.
+	For any $c < 4 \cos(\pi/7)$, \SAT does not have \NAND{} formulas of \textbf{size} $n^{c + o(1)}$.
 \end{corollary}
@@
 
-Uniform \NAND/alternating formulas. Specific model, comparison with the usual model.
+Uniform \NAND{}/alternating formulas. Specific model, comparison with the usual model.
 
-Alternating computation.
+\subsection{Alternating computation}
 
+Given a complexity class $\Cc$ and a constant $c$, we consider \textit{quantified} complexity
+class $(Q~n^c)~\Cc$, where $Q$ is one of $\forall, \exists$ or $\oplus$.
+We define them as follows:
+\begin{definition}[Quantified classes]
+	A language $L$ is in $(Q~n^c)~\Cc$ if and only if there exists a RAM machine
+	that runs in time $n^{c + o(1)}$...
+
+	Problem with this definition: size of the input $n$ is not the same at every stage.
+\end{definition}
+
+\subsection{Overview of the methodology}
 Approach: proof by contradiction.
@@
 \todo[inline]{Prove this}
+Useful starting point : Size-Depth Tradeoffs for Boolean Formulae (Bonet, Buss).
 
+Remark: we must focus on the case when the input formula is a \NAND{} formula.
+Otherwise, we prove a much stronger result.
+Maybe prove the stronger result with a $2+\eps$ factor? Not very useful though.
+If we can go below 2, we improve something.
 %%%%%%%%%%%%%%%%%%%%%%

### 2022-05-19 article/main.tex
Another day

@@
 \newcommand{\Gg}{\mathcal{G}}%
+\newcommand{\Ff}{\mathcal{F}}%
 \newcommand{\Ss}{\mathcal{S}}%
@@
 \newcommand{\NP}{\textsf{NP}}%
+\newcommand{\NTIME}{\textsf{NTIME}}%
 \newcommand{\SAT}{\textsf{SAT}}%
+\newcommand{\ParSAT}{\ensuremath\bigoplus\textsf{SAT}}%
+\newcommand{\SharpSAT}{\#\textsf{SAT}}%
 \newcommand{\NAND}{\textsf{NAND}}%
@@
 
+Slowdown Rule from algorithms for SAT assumptions:
+\begin{lemma}
+	Let $\Ff$ be a class of formulas closed under composition that can simulate projections in constant depth.
+	If $\SAT\in \Ff[c \log n]$, then $\NTIME[n] \subseteq \Ff[c \log n]$.
+\end{lemma}
+\begin{proof}
+	This follows from the fact that $\SAT$ is \NP-complete under uniform poly-log time first-order projections.
+\end{proof}
+
+Can we say the same for $\ParSAT$ and $\SharpSAT$?
+
 How to get the contradiction given the slowdown rule for $c < 4\cos(\pi /7)$.

### 2022-05-26 article/main.tex
Start of article

@@
-\documentclass[a4paper, 12pt]{article}
+\documentclass[a4paper, 11pt]{article}
 % \usepackage{ae,lmodern}
@@
 \usepackage[USenglish]{babel}
+\usepackage[margin=3cm]{geometry}
 \usepackage{amsfonts}
@@
 \newcommand{\NTIME}{\textsf{NTIME}}%
+\newcommand{\coNTIME}{\textsf{coNTIME}}%
 \newcommand{\SAT}{\textsf{SAT}}%
@@
 \newcommand{\NAND}{\textsf{NAND}}%
+\newcommand{\ND}{\textsf{NAND-DEPTH}}%
+\newcommand{\NDL}[1]{\ensuremath\ND[ #1 \log n]}%
 
@@
 	\item $\Omegat{n^3}$ (for both uniform and non-uniform formulas) for Andreev's function \cite{hastad1998shrinkage}\todo{check source} in the De Morgan basis $\{\wedge, \vee, \neg\}$.
-	\item $\Omega{n^2/\log n}$ in the full binary basis (Nechiporuk)
+	\item $\Omega(n^2/\log n)$ in the full binary basis (Nechiporuk)
 	\item $n^{\Theta(\log n)}$ for monotone formulas (formulas over the $\{\wedge, \vee\}$ basis).
@@
 \begin{theorem}
-	For any $c < 4 \cos(\pi/7) \simeq 3.603$, \SAT does not have \NAND{} formulas of depth $(c + o(1)) \log n$.
+	For any $c < 4 \cos(\pi/7) \simeq 3.603$, \SAT{} does not have \NAND{} formulas of depth $(c + o(1)) \log n$.
 \end{theorem}
@@
 \begin{corollary}
-	For any $c < 4 \cos(\pi/7)$, \SAT does not have \NAND{} formulas of \textbf{size} $n^{c + o(1)}$.
+	For any $c < 4 \cos(\pi/7)$, \SAT{} does not have \NAND{} formulas of \textbf{size} $n^{c + o(1)}$.
 \end{corollary}
@@
 \section{Definitions}
-Notations?
+
+\subsection{Uniform formulas}
+A family of formulas
+
+In this work, we focus on formulas with depth $d(n) = (c + o(1))\log n$, for some $c > 1$.
+
+\begin{definition}[``Usual'' polylog-time uniform formulas]
+	A family of formulas $(\varphi_n)_{n \ge 1}$ of depth $(c + o(1))\log n$ 
+	is \textit{polylog-time uniform} if there exists a machine $A_\varphi$
+	such that for every $n$ and every $b\in\bit^{\leq (c + o(1))\log n}$,
+	$A_\varphi(b, n)$ outputs a description of the gate at position $b$ in $\varphi_n$ in time $\log^{O(1)} n$.
+\end{definition}
+
+Here, the position of a gate is given by a bitstring of length at most $(c + o(1))\log n$ 
+that describes the path to follow to reach that gate from the root, 
+with $0$ meaning ``go to left child'' and $1$ ``go to right child''.
+
+In this work, we consider a stronger model of uniform formulas, 
+where the descriptor machine $A_\varphi$ has random access to the input $x$ 
+at which we wish to evaluate the formula, but is still limited to polylogarithmic time.
+\begin{definition}[Input-aware polylog-time uniform formulas]\todo{is this a good name?}\label{def:unif}
+	A family of formulas $(\varphi_n)_{n \ge 1}$ of depth $(c + o(1))\log n$ is \textit{polylog-time uniform}
+	if there exists a \textbf{RAM} machine $B_\varphi$
+	such that for every $n$, every $x\in\bit^n$ and every $b\in\bit^{\leq (c + o(1))\log n}$,
+	$B_\varphi(x, b, n)$ outputs a description of the gate at position $b$ in $\varphi_n$ in time $\log^{O(1)} n$.
+\end{definition}
+Notice that this model supersedes the previous one: 
+$A_\varphi$ satisfies the conditions of Definition~\ref{def:unif}, 
+but does not use its random access to the input, 
+therefore the lower bounds that we give against model also hold for the usual one.
+
+Moreover, this definition allows for an efficient simulation of uniform formulas by RAM machines, 
+due to the negligible overhead.
+Indeed, for every $c > 1$, we have $\ND[c \log n] \subseteq TS[n^c]$.
+
+We will see the need for this stronger definition in \todo{ref}, 
+where we need to give $O(\log n)$ bits of information to the descriptor
+to adaptively choose a subformula.
 
@@
 
-\subsection{Alternating computation}
+\subsection{Quantified/Alternating computation}
 
-Given a complexity class $\Cc$ and a constant $c$, we consider \textit{quantified} complexity
-class $(Q~n^c)~\Cc$, where $Q$ is one of $\forall, \exists$ or $\oplus$.
-We define them as follows:
-\begin{definition}[Quantified classes]
-	A language $L$ is in $(Q~n^c)~\Cc$ if and only if there exists a RAM machine
-	that runs in time $n^{c + o(1)}$...
+We now introduce ``quantified'' complexity classes, which extend alternating classes 
+by adding a ``parity'' quantifier.
 
-	Problem with this definition: size of the input $n$ is not the same at every stage.
+Given a complexity class $\Cc$ and constants $k, (a_i)_{i\leq k}$, 
+we consider \textit{quantified} complexity
+class $(Q_1~n^{a_1})\ldots(Q_k~n^{a_k})~\Cc$, 
+where $Q_i$ is one of $\forall, \exists$ or $\oplus$.
+A computation in a quantified complexity class is defined as follows:
+\begin{definition}[Quantified computation]
+	A language $L$ is in $(Q_1~n^{a_1})\ldots(Q_k~n^{a_k})~\Cc$ 
+	is defined by $k$ RAM machines $A_1, \ldots, A_k$ and a $\Cc$ machine $M$
+	such that, given an input $x$ of length $n$:
+	for every $i \le k$, $A_{i}$ takes as input $x_i$ along with $y_i$, 
+	a string of $n^{a_{i} + o(1)}$ nondeterministic bits,
+	runs in time $n^{a_{i} + o(1)}$ and outputs a string $x_{i+1}$ of length at most $n^{a_{i} + o(1)}$, 
+	with the convention that $x_0 := x$.
+	The output of the last stage, $x_{k+1}$, is passed to the machine $M$.
+
+	The computation accepts $x$ \textit{starting from stage $i$} if either:
+	\begin{itemize}
+		\item $i = k+1$ and $M$ accepts $x_{k+1}$, 
+		\item $Q_i$ is $\exists$, and there exists a $y_i$ such that the computation
+		accepts $x_{i+1}$ starting from stage $i+1$, 
+		\item $Q_i$ is $\forall$, and for every $y_i$, the computation
+		accepts $x_{i+1}$ starting from stage $i+1$, 
+		\item $Q_i$ is $\bigoplus$, the number of $y_i$ such that the computation
+		accepts $x_{i+1}$ starting from stage $i+1$ is odd. 
+	\end{itemize}
+
+	An input $x$ is in $L$ if the computation $x$ accepts starting from stage 1.
 \end{definition}
@@
 \subsection{Overview of the methodology}
-Approach: proof by contradiction.
+
+The general methodology used in this paper to obtain lower bounds
+extends the one used in~\cite{mudigonda2020time}.
+
+To prove that \SAT{} does not have formulas of depth less than $c\log n$, 
+we reason by contradiction.
+First, we assume that \SAT{} has such formulas, and the strong \NP-completeness of \SAT{}
+allows us to deduce that the same is holds for all of $\NTIME[n]$, 
+a result referred to as the ``Slowdown Lemma''.
+\begin{lemma}[Slowdown Lemma]
+	Let $\Ff$ be an expressive\todo{define} class of formulas, and let $c > 1$.
+	If \SAT{} is in $\Ff[c \log n]$\todo{define},
+	then $\NTIME[n] \cup \coNTIME[n] \subseteq \Ff[c \log n]$.
+\end{lemma}
+\todo[inline]{talk about the issue with input size}
+
+We also show how to speed-up (or rather, reduce the depth of) computations by adding quantifiers, 
+using various ``Speedup Lemmas''. 
+For example, for NAND formulas (which are the same as alternating AND/OR formulas), we have the following:
+\begin{lemma}[Speedup Lemma]
+	For every $c > 0, x \leq c$, we have:
+	\[\textsf{NAND-depth}[c \log n] \subseteq (\exists n^{x/2}) (\forall n^1) \textsf{NAND-depth}[(c-x) \log n]\]
+\end{lemma}
+
+We then combine multiple application of these lemmas with various parameters,
+with the goal of contradicting a known hierarchy theorem.
+For example, we can prove that SAT does not have NAND formulas 
+of depth less than $(2-\eps) \log n$ for any $\eps > 0$ with the following proof.
+Assume that $\SAT\in \textsf{NAND-depth}[c \log n]$.
+Then, we have
+\begin{flalign*}
+	NTIME[n^{2}]
+		&\subseteq \NDL{2c}	& \text{Slowdown Lemma}\\
+		&\subseteq (\exists n^{c})(\forall n^1)\NDL{0}& \text{Speedup Lemma with param. } x = 2c\\
+		&\subseteq (\exists n^{c})\coNTIME[n]	& \\
+		&\subseteq (\exists n^{c})\NDL{c}	& \text{Slowdown Lemma}\\
+		&\subseteq \NTIME[n^c]	&
+\end{flalign*}
+For $c < 2$, this contradicts the Nondeterministic Time Hierarchy Theorem.
+In practice, we use (integer) linear programming to find the best order of application
+of the rules, and the best parameters to use.
+
+We proceed similarly with properties of \ParSAT{} and \SharpSAT{}
+to prove results for fomulas with XOR gates.
 
@@
 	Let $\Ff$ be a class of formulas closed under composition that can simulate projections in constant depth.
-	If $\SAT\in \Ff[c \log n]$, then $\NTIME[n] \subseteq \Ff[c \log n]$.
+	If $\SAT{}\in \Ff[c \log n]$, then $\NTIME[n] \subseteq \Ff[c \log n]$.
 \end{lemma}
 \begin{proof}
-	This follows from the fact that $\SAT$ is \NP-complete under uniform poly-log time first-order projections.
+	This follows from the fact that $\SAT{}$ is \NP-complete under uniform poly-log time first-order projections.
 \end{proof}

### 2022-05-26 article/main.tex
Discussion of speedup and slowdown

@@
 \newcommand{\NAND}{\textsf{NAND}}%
-\newcommand{\ND}{\textsf{NAND-DEPTH}}%
+\newcommand{\ND}{\textsf{NDepth}}%
 \newcommand{\NDL}[1]{\ensuremath\ND[ #1 \log n]}%
@@
 	runs in time $n^{a_{i} + o(1)}$ and outputs a string $x_{i+1}$ of length at most $n^{a_{i} + o(1)}$, 
-	with the convention that $x_0 := x$.
+	with the convention that $x_1 := x$.
 	The output of the last stage, $x_{k+1}$, is passed to the machine $M$.
@@
 \end{definition}
+\todo[inline]{allow $O(\log n)$ quantifiers ?}
 
@@
 allows us to deduce that the same is holds for all of $\NTIME[n]$, 
-a result referred to as the ``Slowdown Lemma''.
-\begin{lemma}[Slowdown Lemma]
+a result referred to as the ``Slowdown rule''.
+\begin{lemma}[Slowdown rule]
 	Let $\Ff$ be an expressive\todo{define} class of formulas, and let $c > 1$.
@@
 We also show how to speed-up (or rather, reduce the depth of) computations by adding quantifiers, 
-using various ``Speedup Lemmas''. 
+using various ``Speedup rules''. 
 For example, for NAND formulas (which are the same as alternating AND/OR formulas), we have the following:
-\begin{lemma}[Speedup Lemma]
+\begin{lemma}[Speedup rule]
 	For every $c > 0, x \leq c$, we have:
@@
 	NTIME[n^{2}]
-		&\subseteq \NDL{2c}	& \text{Slowdown Lemma}\\
-		&\subseteq (\exists n^{c})(\forall n^1)\NDL{0}& \text{Speedup Lemma with param. } x = 2c\\
+		&\subseteq \NDL{2c}	& \text{Slowdown rule}\\
+		&\subseteq (\exists n^{c})(\forall n^1)\NDL{0}& \text{Speedup rule with param. } x = 2c\\
 		&\subseteq (\exists n^{c})\coNTIME[n]	& \\
-		&\subseteq (\exists n^{c})\NDL{c}	& \text{Slowdown Lemma}\\
+		&\subseteq (\exists n^{c})\NDL{c}	& \text{Slowdown rule}\\
 		&\subseteq \NTIME[n^c]	&
@@
 For $c < 2$, this contradicts the Nondeterministic Time Hierarchy Theorem.
+
 In practice, we use (integer) linear programming to find the best order of application
@@
 
-We proceed similarly with properties of \ParSAT{} and \SharpSAT{}
-to prove results for fomulas with XOR gates.
-
 %%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%
-\section{Proof of thm}
-Speedup Rule for Alternating formulas.
+\section{Proof of main result}
 
-Slowdown Rule from algorithms for SAT assumptions:
-\begin{lemma}
-	Let $\Ff$ be a class of formulas closed under composition that can simulate projections in constant depth.
-	If $\SAT{}\in \Ff[c \log n]$, then $\NTIME[n] \subseteq \Ff[c \log n]$.
+\subsection{Speedup and Slowdowns}
+We start by Slowdown rule from algorithms for SAT assumptions:
+\begin{lemma}[Slowdown rule]\label{lemma:slowdown}
+	If $\SAT{}\in \NDL{c}$, then for every $d > 1$, $\NTIME[n^d] \cup \coNTIME[n^d] \subseteq \NDL{c\cdot d}$.
 \end{lemma}
-\begin{proof}
-	This follows from the fact that $\SAT{}$ is \NP-complete under uniform poly-log time first-order projections.
+
+A key idea for the proof of Lemma~\ref{lemma:slowdown} 
+is the fact that any language in $\NTIME[n]$ can
+be reduced to $\SAT{}$ via uniform poly-log time first-order projections, 
+with instances of size $n$ reduced to formulas of size $n^{1+o(1)}$.
+Intuitively, a first order projection is a depth 0 circuit, 
+i.e. each output bit is either a constant (0 or 1),
+an input $x_i$ or its negation, $\neg x_i$.
+For a more in-depth exposition of first-order projections,
+see for example~\cite[End of Sec.~3]{allender1997first}.
+\begin{proof}[Proof of Lemma~\ref{lemma:slowdown}]
+	We first prove that $\SAT{}\in \NDL{c}$ implies that $\NTIME[n] \subseteq \NDL{c}$,
+	then use a padding argument to show that $\forall d>1, \NTIME[n^d] \subseteq \NDL{cd}$,
+	and the result follows since $\NDL{cd}$ is closed under complement.
+
+	Let $L\in\NTIME[n]$. 
+	As stated above, $L$ is Karp-reducible to \SAT{} via
+	uniform poly-log time first-order projections, 
+	\todo{ref for this? Folklore?}
+	with instances of size $n$ reduced to formulas of size $n^{1+o(1)}$.
+	Now, notice that the composition of a NAND formula of depth $d$ with a first order projection
+	is also a NAND formula of depth $d$: it only consists in relabeling leaves of the formula.
+	Poly-log time uniformity is also preserved. 
+	Therefore, we can use $(c+o(1)) \log n$-depth NAND formulas for \SAT{} on $n^{1+o(1)}$
+	inputs to decide $L$ by composing them with the first order projection; 
+	the resulting formula has size $(c+o(1)) \log n^{1+o(1)} = (c+o(1)) \log n$,
+	therefore $L\in\NDL{c}$.
+
+	We now use a padding argument to lift this result to $\NTIME[n^d]$.
+	Let $d>1$, let $L\in \NTIME[n^d]$ and let $M$ be a nondeterministic machine that decides $L$
+	in time $n^{d+o(1)}$.
+	Consider the padded language $L' = \{x1^{|x|^d - |x|} \mid x\in L\}$.
+	Given an input $y$ of length $m$, we can decide whether $y\in L'$ in time 
+	$m^{1+o(1)}$ as follows:
+	\begin{itemize}
+		\item if $y$ is not of the form $x1^{|x|^d - |x|}$ for some $x$, reject $y$.
+		This can be done in time $O(m)$.
+		\item otherwise, $y = x1^{|x|^d - |x|}$. Run $M$ on $x$, and accept $y$ 
+		if and only it accepts $x$. This takes time $|x|^{d+o(1)} = m^{1+o(1)}$.
+	\end{itemize}
+	This implies $L'$ is in $\NTIME[n]$, which in turn implies that $L'\in\NDL{c}$.
+	We can then use the formula descriptor $A$ for $L'$ 
+	to describe formulas of depth $(c + o(1))\log \left(n^{d+o(1)}\right) = (cd + o(1))\log n$ for $L$.
+	On input $x,b,n$, the descriptor machine $B$ for $L$  
+	calls $A$ on $x, b, n^d$.
+	If $A$ output a gate type or a constant, $B$ outputs the same information.
+	If $A$ outputs a literal $x_i^*$, $B$ outputs $x_i^*$ if $i \le n$, and 1 otherwise.
 \end{proof}
 
-Can we say the same for $\ParSAT$ and $\SharpSAT$?
+We need to be extra careful when applying the slowdown lemma to alternating classes:
+the $i$-th computation stage takes as input a string of length $m = n^{a_i}$, not $n$.
+Since the previous result only holds for $\NTIME[n^d], d > 1$, 
+we need to take into
+account the edge case where the running time of the $i$-th
+stage is sublinear in the size of its input.
+We have the following:
+\begin{corollary}[Extended Slowdown rule]
+	If $\SAT{}\in \NDL{c}$, then for every $k \in\NN$ and $d > 1$, we have:
+	\[\ldots (Q~n^{a}) (\neg Q~n^b) \NDL{d}
+	\subseteq \ldots (Q~n^{a}) (\neg Q~n^b) \NDL{c\cdot\max(d, a, b)}.\]
+\end{corollary}
+\begin{proof}
+	TODO
+\end{proof}
+
+We now discuss
+Speedup rule for Alternating formulas.
+
+Intuitively, the only information needed to know that an OR gate evaluates
+to 1 is that one of its children evaluates to 1.
+The role of the existential quantifier is to guess for every OR gate in the first $(x/2)\log n$ levels,
+which of its children evaluates to 1, and we only evaluate the corresponding subformula.
+Moreover, making this guess reduces the number of OR gates that we will evaluate
+(this number doubles every two levels, instead of doubling every level), 
+therefore we only need to guess $n^{x/2}$ bits. 
+We can then eliminate OR gates and replace them with their selected child.
+We are left with a formula starting with $(x/2)\log n$ levels, 
+followed by an OR/AND formula of depth $(c-x + o(1)) \log n$.
+The first part can be considered as an AND gate on $n^{x/2}$ inputs (the smaller formulas).
+This formula evaluate to 1 if and only if for every of these $n^{x/2}$
+smaller formulas evaluates to true: we can check this using a universal quantifiers, 
+this requires $O(\frac{x}{2}\log n)$ bits.
 

### 2022-06-10 article/main.tex
Started Sym of Sym fast TT eval report

@@
 \newcommand{\NP}{\textsf{NP}}%
+\newcommand{\TIME}{\textsf{TIME}}%
 \newcommand{\NTIME}{\textsf{NTIME}}%
@@
 \newcommand{\ND}{\textsf{NDepth}}%
-\newcommand{\NDL}[1]{\ensuremath\ND[ #1 \log n]}%
+\newcommand{\NDL}[1]{\ND[ #1 \log n]}%
+
+\newcommand{\todohere}{\todo[inline]{TODO}}%
 
@@
 	a string of $n^{a_{i} + o(1)}$ nondeterministic bits,
-	runs in time $n^{a_{i} + o(1)}$ and outputs a string $x_{i+1}$ of length at most $n^{a_{i} + o(1)}$, 
-	with the convention that $x_1 := x$.
-	The output of the last stage, $x_{k+1}$, is passed to the machine $M$.
+	runs in time $n^{a_{i} + o(1)}$ and outputs a string $x_{i}$ of length at most $n^{a_{i} + o(1)}$, 
+	with the convention that $x_0 := x$.
+	The output of the last stage, $x_{k}$, is passed to the machine $M$.
 
@@
 	\begin{itemize}
-		\item $i = k+1$ and $M$ accepts $x_{k+1}$, 
+		\item $i = k+1$ and $M$ accepts $x_{k}$, 
 		\item $Q_i$ is $\exists$, and there exists a $y_i$ such that the computation
-		accepts $x_{i+1}$ starting from stage $i+1$, 
+		accepts $x_{i}$ starting from stage $i+1$, 
 		\item $Q_i$ is $\forall$, and for every $y_i$, the computation
-		accepts $x_{i+1}$ starting from stage $i+1$, 
+		accepts $x_{i}$ starting from stage $i+1$, 
 		\item $Q_i$ is $\bigoplus$, the number of $y_i$ such that the computation
-		accepts $x_{i+1}$ starting from stage $i+1$ is odd. 
+		accepts $x_{i}$ starting from stage $i+1$ is odd. 
 	\end{itemize}
@@
 \begin{lemma}[Slowdown rule]
-	Let $\Ff$ be an expressive\todo{define} class of formulas, and let $c > 1$.
-	If \SAT{} is in $\Ff[c \log n]$\todo{define},
-	then $\NTIME[n] \cup \coNTIME[n] \subseteq \Ff[c \log n]$.
+	% Let $\Ff$ be an expressive\todo{define} class of formulas, and let $c > 1$.
+	If \SAT{} is in $\NDL{c}$ for some $c>0$,
+	then $\NTIME[n] \cup \coNTIME[n] \subseteq \NDL{c}$.
 \end{lemma}
-\todo[inline]{talk about the issue with input size}
+\todo[inline]{That's not the rule...}
 
@@
 with the goal of contradicting a known hierarchy theorem.
+Assuming that $\NTIME[n] \subseteq \NDL[c]$ for some $n$, we obtain the following hierarchy theorem
+for NAND formulas, as a corollary of the Nondeterministic Time Hierarchy Theorem:
+\begin{theorem}\label{thm:nandh}
+	If there exists $c > 0$ such that $\NTIME[n] \subseteq \NDL{c}$, 
+	then for all $b > a > 0$, we have
+	\[\NDL{a} \subsetneq \NDL{b}.\] 
+\end{theorem}
+
 For example, we can prove that SAT does not have NAND formulas 
-of depth less than $(2-\eps) \log n$ for any $\eps > 0$ with the following proof.
-Assume that $\SAT\in \textsf{NAND-depth}[c \log n]$.
+of depth less than $2.8284 \log n$ with the following proof.
+Assume that $\SAT\in \textsf{NAND-depth}[c \log n]$ for some $c > 0$.
 Then, we have
 \begin{flalign*}
-	NTIME[n^{2}]
-		&\subseteq \NDL{2c}	& \text{Slowdown rule}\\
-		&\subseteq (\exists n^{c})(\forall n^1)\NDL{0}& \text{Speedup rule with param. } x = 2c\\
-		&\subseteq (\exists n^{c})\coNTIME[n]	& \\
-		&\subseteq (\exists n^{c})\NDL{c}	& \text{Slowdown rule}\\
-		&\subseteq \NTIME[n^c]	&
+	\NDL{4}
+		&\subseteq (\exists n)(\forall n)\NDL{2}& \text{Speedup rule with param. } x = 2\\
+		&\subseteq (\exists n)\NDL{c}	& \text{Slowdown rule}\\
+		&\subseteq \NDL{c^2/2}	& \text{Slowdown rule}
 \end{flalign*}
-For $c < 2$, this contradicts the Nondeterministic Time Hierarchy Theorem.
+For $c^2/2  < 4 \leftrightarrow c < 2\sqrt{2} \simeq 2.8284\ldots$, this contradicts Theorem~\ref{thm:nandh}.
 
@@
 \subsection{Speedup and Slowdowns}
-We start by Slowdown rule from algorithms for SAT assumptions:
+We start by discussion the Speedup rule for Alternating formulas.
+
+\begin{lemma}[Speedup rule]\label{lemma:speedup}
+	For every $d > 0$ and $0 \ge x \ge d$, we have
+	\begin{flalign*}
+	\NDL{d} &\subseteq (\exists n^{x/2}) (\forall \frac{x}{2}\log n) \NDL{d-x}\text{ and}\\
+	\NDL{d} &\subseteq (\forall n^{x/2}) (\exists \frac{x}{2}\log n) \NDL{d-x}
+	\end{flalign*}
+\end{lemma}
+
+The intuitive idea behind this lemma is that the only information needed to know that an OR gate evaluates
+to 1 is that one of its children evaluates to 1. 
+We apply this idea to every OR gate in the first $x \log n$ levels of the formula,
+and then use a universal quantifier to handle the remaining AND gates.
+
+The role of the existential quantifier is to guess for every OR gate in the first $(x/2)\log n$ levels,
+which of its children evaluates to 1, and we only evaluate the corresponding subformula.
+Moreover, making this guess reduces the number of OR gates that we will evaluate
+(this number doubles every two levels, instead of doubling at every level), 
+therefore we only need to guess $O(n^{x/2})$ bits.
+
+We can then eliminate OR gates and replace them with their selected child.
+We are left with a formula starting with $(x/2)\log n$ levels of AND gates, 
+followed by OR/AND formulas of depth $(d-x + o(1)) \log n$.
+The main formula evaluate to 1 if and only if for every of these $O(n^{x/2})$
+smaller formulas evaluates to true: 
+we can check this using a universal quantifiers, 
+which requires $O(\frac{x}{2}\log n)$ bits, and then running an $\NDL{d-x}$ computation.
+
+Finally, notice that the definition of alternating computation
+requires that there exists a \textit{single} machine $M$ that runs the
+final stage of computation \textit{for every} possible choice of nondeterministic bits.
+This is where we use the strong notion of uniformity (from Definition~\ref{def:unif}), where the descriptor machine
+has random access to the input: giving $M$ access to (a few) bits of the input
+allows us to adapt its behavior depending on the nondeterministic bits.
+
+\begin{proof}[Proof of Lemma~\ref{lemma:speedup}]
+	We prove the result for $\exists\forall$ computations, 
+	the result for $\forall\exists$ computations follows from the fact that $\NDL{d}$ is closed
+	under complementation.
+
+	Let $L\in\NDL{d}$, let be $M : (x,i,n) \mapsto g$ be the a descriptor machine for $L$,
+	and let $x$ be an input of length $n$.
+
+	We now construct a machine $M'$ that describes $\NDL{d-x}$-formulas,
+	such that an $\exists\forall$ computation with $M'$ as its last stage accepts $L$.
+	On an input $x$
+
+	We describe how to build bit strings $y$ of length $n^{x/2}$ and
+	\todohere 
+\end{proof}
+
+\begin{corollary}\label{cor:speedup}
+	For every $d > 0$, we have
+	\[\NDL{d} \subseteq \NTIME[n^{d/2}] \cap \coNTIME[n^{d/2}].\]
+\end{corollary}
+\begin{proof}
+	Applying Lemma~\ref{lemma:speedup} with $x = d$, we obtain
+	$\NDL{d} \subseteq (\exists n^{d/2}) (\forall \frac{d}{2}\log n) \NDL{0}$.
+	Now, using an exhaustive search over the $n^{d/2}$ bit strings of length $\frac{d}{2}\log n$
+	followed by a deterministic $n^{o(1)}$ computation,
+	we get
+	$\NDL{d} \subseteq (\exists n^{d/2}) \TIME[n^{d/2}]$.
+\end{proof}
+
+We now discuss
+Slowdown rule from algorithms for SAT assumptions:
 \begin{lemma}[Slowdown rule]\label{lemma:slowdown}
-	If $\SAT{}\in \NDL{c}$, then for every $d > 1$, $\NTIME[n^d] \cup \coNTIME[n^d] \subseteq \NDL{c\cdot d}$.
+	If $\SAT{}\in \NDL{c}$, then for every $d \ge 1$, $\NTIME[n^d] \cup \coNTIME[n^d] \subseteq \NDL{c\cdot d}$.
 \end{lemma}
@@
 	We first prove that $\SAT{}\in \NDL{c}$ implies that $\NTIME[n] \subseteq \NDL{c}$,
-	then use a padding argument to show that $\forall d>1, \NTIME[n^d] \subseteq \NDL{cd}$,
-	and the result follows since $\NDL{cd}$ is closed under complement.
+	then use a padding argument to show that $\forall d\ge 1, \NTIME[n^d] \subseteq \NDL{c\cdot d}$,
+	and the result follows since $\NDL{c\cdot d}$ is closed under complement.
 
@@
 	We can then use the formula descriptor $A$ for $L'$ 
-	to describe formulas of depth $(c + o(1))\log \left(n^{d+o(1)}\right) = (cd + o(1))\log n$ for $L$.
+	to describe formulas of depth $(c + o(1))\log \left(n^{d+o(1)}\right) = (c\cdot d + o(1))\log n$ for $L$.
 	On input $x,b,n$, the descriptor machine $B$ for $L$  
@@
 
-We need to be extra careful when applying the slowdown lemma to alternating classes:
-the $i$-th computation stage takes as input a string of length $m = n^{a_i}$, not $n$.
-Since the previous result only holds for $\NTIME[n^d], d > 1$, 
+We need to be careful when applying the slowdown lemma to alternating classes:
+the $i+1$-th computation stage takes as input a string of length $m = n^{a_i}$, not $n$.
+Since the previous result only holds for $\NTIME[n^d], d \ge 1$, 
 we need to take into
-account the edge case where the running time of the $i$-th
+account the edge case where the running time of the $i+1$-th
 stage is sublinear in the size of its input.
@@
 \begin{corollary}[Extended Slowdown rule]
-	If $\SAT{}\in \NDL{c}$, then for every $k \in\NN$ and $d > 1$, we have:
+	If $\SAT{}\in \NDL{c}$, then for every $d > 1$, we have:
 	\[\ldots (Q~n^{a}) (\neg Q~n^b) \NDL{d}
-	\subseteq \ldots (Q~n^{a}) (\neg Q~n^b) \NDL{c\cdot\max(d, a, b)}.\]
+	\subseteq \ldots (Q~n^{a}) \NDL{c\cdot\max(d/2, a, b)}.\]
+	\todo{fix the $d/2$}
 \end{corollary}
 \begin{proof}
-	TODO
+	Without loss of generality, assume that $Q = \forall$, and therefore $\neg Q = \exists$.
+	In that case, we have $(\neg Q~n^b) \NDL{d} \subseteq \NTIME[n^{\max(d, b)}]$,
+	since evaluating a formula of depth $d \log n$ can be done in deterministic time
+	$n^{d + o(1)}$.
+
+	Now, consider the $\NTIME[n^{\max(d, b)}]$ language $L$ that correspond to the last stage of the alternating 
+	computation, and its corresponding nondeterministic machine $M$. 
+	It takes as input a string of length $m = n^{a}$, and its runtime, as a function of $m$, 
+	is $m^{\max(d, b)/a}$.
+	We apply the Slowdown rule to $M$. There are two possible cases:
+	\begin{itemize}
+		\item if $\max(d, b)/a < 1$, then we have $L \in \NTIME[m^{\max(d, b)/a}] \subseteq \NTIME[m]
+		\subseteq \ND[c \log m] \subseteq \NDL{c \cdot a}$.
+		Since  $\max(d, b) < a$, we have $a = \max(d, a, b)$ and therefore 
+		$\NDL{c \cdot a} = \NDL{c\cdot\max(d, a, b)}$.
+		\item otherwise, $\max(d, b) \ge a$, and applying the slowdown rule yields
+		$L \in \NTIME[m^{\max(d, b)/a}] =  \NTIME[n^{\max(d, b)}] \subseteq \NDL{c \cdot \max(d, a, b)}$.
+	\end{itemize}
 \end{proof}
-
-We now discuss
-Speedup rule for Alternating formulas.
-
-Intuitively, the only information needed to know that an OR gate evaluates
-to 1 is that one of its children evaluates to 1.
-The role of the existential quantifier is to guess for every OR gate in the first $(x/2)\log n$ levels,
-which of its children evaluates to 1, and we only evaluate the corresponding subformula.
-Moreover, making this guess reduces the number of OR gates that we will evaluate
-(this number doubles every two levels, instead of doubling every level), 
-therefore we only need to guess $n^{x/2}$ bits. 
-We can then eliminate OR gates and replace them with their selected child.
-We are left with a formula starting with $(x/2)\log n$ levels, 
-followed by an OR/AND formula of depth $(c-x + o(1)) \log n$.
-The first part can be considered as an AND gate on $n^{x/2}$ inputs (the smaller formulas).
-This formula evaluate to 1 if and only if for every of these $n^{x/2}$
-smaller formulas evaluates to true: we can check this using a universal quantifiers, 
-this requires $O(\frac{x}{2}\log n)$ bits.
 

### 2022-06-10 sym_eval/main.tex
Started Sym of Sym fast TT eval report

--- 2022-05-26 article/main.tex
@@
 \newcommand{\NAND}{\textsf{NAND}}%
-\newcommand{\ND}{\textsf{NAND-DEPTH}}%
-\newcommand{\NDL}[1]{\ensuremath\ND[ #1 \log n]}%
+\newcommand{\ND}{\textsf{NDepth}}%
+\newcommand{\NDL}[1]{\ND[ #1 \log n]}%
+\newcommand{\ssym}{\textsf{SYM$\circ$SYM}}%
+
+\newcommand{\todohere}{\todo[inline]{TODO}}%
 
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\title{Uniform formula lower bounds for finding and counting satisfiable assignments}
+\title{Fast multipoint evaluation for exponential size CNFs and SYM$\circ$SYM circuits}
 \author{Gabriel Bathie, Ryan Williams}
@@
 \begin{abstract}
+We introduce a new way to encode size-$m$ \ssym{} circuits with possibly negated inputs
+(which include CNFs and DNFs)
+into polynomials that can be evaluated over all inputs in $\bit^n$ in time $O^*(2^n + m)$.
+As a consequence, we obtain a $O^*(2^n)$-time algorithm that computes the
+truthe table of $2^n$-size CNFs.
+Lower bounds ensue. (do they though?)
 \end{abstract}
-
-List of things to do:
-\begin{itemize}
-	\item Improve Spira/Brent to show that a $c \log n$ depth 
-	lower bound implies an $n^{c-\eps}$ size lower bound?
-	\item Show that this bound is optimal w.r.t. these rules?
-\end{itemize}
 
@@
 
-Motivation: understanding complexity. Explicit lower bounds : programs, circuits, formulas.
-Uniformity.
-
-Related work:
-- Original line of work : \cite{fortnow2000time,fortnow2005time}
-- Main line of work \cite{williams2006inductive,williams2007time,williams2013alternation} and its dramatic resolution \cite{buss2015limits}.
-- Broader classes \cite{mudigonda2020time}
-
-Best known formula lower bounds for languages (functions?) in NP: 
-\begin{itemize}
-	\item $\Omegat{n^3}$ (for both uniform and non-uniform formulas) for Andreev's function \cite{hastad1998shrinkage}\todo{check source} in the De Morgan basis $\{\wedge, \vee, \neg\}$.
-	\item $\Omega(n^2/\log n)$ in the full binary basis (Nechiporuk)
-	\item $n^{\Theta(\log n)}$ for monotone formulas (formulas over the $\{\wedge, \vee\}$ basis).
-\end{itemize}
-
 \subsection{Results overview}
 
-We first study \NAND{} formulas. 
-The \NAND{} basis is rather expressive: Andreev's function has \NAND{} formulas of depth $(3+o(1)) \log n$ (and therefore of size $\Ot{n^3}$), which matches the lower bound for the De Morgan Basis.
-
 \begin{theorem}
-	For any $c < 4 \cos(\pi/7) \simeq 3.603$, \SAT{} does not have \NAND{} formulas of depth $(c + o(1)) \log n$.
+	For all $n, m$, \ssym{} circuits of size $m$ over $n$ variables can be evaluated
+	on every input in time $O^*(2^n + m)$.
 \end{theorem}
 
-\todo[inline]{Prove all of the following.
-
-After checking \cite{sergeev2019relation}, this is probably incorrect...}
-\begin{lemma}
-	For any \NAND{} formula on $n$ inputs of size $n^c$ and any $\eps > 0$, 
-	there exists an equivalent \NAND{} formula of depth $(c+\eps) \log n$.
-\end{lemma}
+This yields a significant improvement over the exhaustive search algorithm, that takes time $O(2^nm)$.
 
@@
 
-\subsection{Uniform formulas}
-A family of formulas
-
-In this work, we focus on formulas with depth $d(n) = (c + o(1))\log n$, for some $c > 1$.
-
-\begin{definition}[``Usual'' polylog-time uniform formulas]
-	A family of formulas $(\varphi_n)_{n \ge 1}$ of depth $(c + o(1))\log n$ 
-	is \textit{polylog-time uniform} if there exists a machine $A_\varphi$
-	such that for every $n$ and every $b\in\bit^{\leq (c + o(1))\log n}$,
-	$A_\varphi(b, n)$ outputs a description of the gate at position $b$ in $\varphi_n$ in time $\log^{O(1)} n$.
+A symmetric function is a boolean function whose output only depends on the number of 
+its input that equal to 1.
+More formally:
+\begin{definition}[Symmetric functions]
+	A boolean function $f: \bit^n \mapsto \bit$ is \textit{symmetric} if there exists a function
+	$g: \{0,\ldots, n\} \mapsto \bit$ such that for every $x\bit^n$:
+	\[f(x) = g(\sum_{i=1}^n x_i).\]
 \end{definition}
 
-Here, the position of a gate is given by a bitstring of length at most $(c + o(1))\log n$ 
-that describes the path to follow to reach that gate from the root, 
-with $0$ meaning ``go to left child'' and $1$ ``go to right child''.
-
-In this work, we consider a stronger model of uniform formulas, 
-where the descriptor machine $A_\varphi$ has random access to the input $x$ 
-at which we wish to evaluate the formula, but is still limited to polylogarithmic time.
-\begin{definition}[Input-aware polylog-time uniform formulas]\todo{is this a good name?}\label{def:unif}
-	A family of formulas $(\varphi_n)_{n \ge 1}$ of depth $(c + o(1))\log n$ is \textit{polylog-time uniform}
-	if there exists a \textbf{RAM} machine $B_\varphi$
-	such that for every $n$, every $x\in\bit^n$ and every $b\in\bit^{\leq (c + o(1))\log n}$,
-	$B_\varphi(x, b, n)$ outputs a description of the gate at position $b$ in $\varphi_n$ in time $\log^{O(1)} n$.
-\end{definition}
-Notice that this model supersedes the previous one: 
-$A_\varphi$ satisfies the conditions of Definition~\ref{def:unif}, 
-but does not use its random access to the input, 
-therefore the lower bounds that we give against model also hold for the usual one.
-
-Moreover, this definition allows for an efficient simulation of uniform formulas by RAM machines, 
-due to the negligible overhead.
-Indeed, for every $c > 1$, we have $\ND[c \log n] \subseteq TS[n^c]$.
-
-We will see the need for this stronger definition in \todo{ref}, 
-where we need to give $O(\log n)$ bits of information to the descriptor
-to adaptively choose a subformula.
-
-Uniform \NAND{}/alternating formulas. Specific model, comparison with the usual model.
-
-\subsection{Quantified/Alternating computation}
-
-We now introduce ``quantified'' complexity classes, which extend alternating classes 
-by adding a ``parity'' quantifier.
-
-Given a complexity class $\Cc$ and constants $k, (a_i)_{i\leq k}$, 
-we consider \textit{quantified} complexity
-class $(Q_1~n^{a_1})\ldots(Q_k~n^{a_k})~\Cc$, 
-where $Q_i$ is one of $\forall, \exists$ or $\oplus$.
-A computation in a quantified complexity class is defined as follows:
-\begin{definition}[Quantified computation]
-	A language $L$ is in $(Q_1~n^{a_1})\ldots(Q_k~n^{a_k})~\Cc$ 
-	is defined by $k$ RAM machines $A_1, \ldots, A_k$ and a $\Cc$ machine $M$
-	such that, given an input $x$ of length $n$:
-	for every $i \le k$, $A_{i}$ takes as input $x_i$ along with $y_i$, 
-	a string of $n^{a_{i} + o(1)}$ nondeterministic bits,
-	runs in time $n^{a_{i} + o(1)}$ and outputs a string $x_{i+1}$ of length at most $n^{a_{i} + o(1)}$, 
-	with the convention that $x_0 := x$.
-	The output of the last stage, $x_{k+1}$, is passed to the machine $M$.
-
-	The computation accepts $x$ \textit{starting from stage $i$} if either:
-	\begin{itemize}
-		\item $i = k+1$ and $M$ accepts $x_{k+1}$, 
-		\item $Q_i$ is $\exists$, and there exists a $y_i$ such that the computation
-		accepts $x_{i+1}$ starting from stage $i+1$, 
-		\item $Q_i$ is $\forall$, and for every $y_i$, the computation
-		accepts $x_{i+1}$ starting from stage $i+1$, 
-		\item $Q_i$ is $\bigoplus$, the number of $y_i$ such that the computation
-		accepts $x_{i+1}$ starting from stage $i+1$ is odd. 
-	\end{itemize}
-
-	An input $x$ is in $L$ if the computation $x$ accepts starting from stage 1.
+\begin{definition}[\ssym{} circuits]
 \end{definition}
 
-\subsection{Overview of the methodology}
-
-The general methodology used in this paper to obtain lower bounds
-extends the one used in~\cite{mudigonda2020time}.
-
-To prove that \SAT{} does not have formulas of depth less than $c\log n$, 
-we reason by contradiction.
-First, we assume that \SAT{} has such formulas, and the strong \NP-completeness of \SAT{}
-allows us to deduce that the same is holds for all of $\NTIME[n]$, 
-a result referred to as the ``Slowdown Lemma''.
-\begin{lemma}[Slowdown Lemma]
-	Let $\Ff$ be an expressive\todo{define} class of formulas, and let $c > 1$.
-	If \SAT{} is in $\Ff[c \log n]$\todo{define},
-	then $\NTIME[n] \cup \coNTIME[n] \subseteq \Ff[c \log n]$.
-\end{lemma}
-\todo[inline]{talk about the issue with input size}
-
-We also show how to speed-up (or rather, reduce the depth of) computations by adding quantifiers, 
-using various ``Speedup Lemmas''. 
-For example, for NAND formulas (which are the same as alternating AND/OR formulas), we have the following:
-\begin{lemma}[Speedup Lemma]
-	For every $c > 0, x \leq c$, we have:
-	\[\textsf{NAND-depth}[c \log n] \subseteq (\exists n^{x/2}) (\forall n^1) \textsf{NAND-depth}[(c-x) \log n]\]
-\end{lemma}
-
-We then combine multiple application of these lemmas with various parameters,
-with the goal of contradicting a known hierarchy theorem.
-For example, we can prove that SAT does not have NAND formulas 
-of depth less than $(2-\eps) \log n$ for any $\eps > 0$ with the following proof.
-Assume that $\SAT\in \textsf{NAND-depth}[c \log n]$.
-Then, we have
-\begin{flalign*}
-	NTIME[n^{2}]
-		&\subseteq \NDL{2c}	& \text{Slowdown Lemma}\\
-		&\subseteq (\exists n^{c})(\forall n^1)\NDL{0}& \text{Speedup Lemma with param. } x = 2c\\
-		&\subseteq (\exists n^{c})\coNTIME[n]	& \\
-		&\subseteq (\exists n^{c})\NDL{c}	& \text{Slowdown Lemma}\\
-		&\subseteq \NTIME[n^c]	&
-\end{flalign*}
-For $c < 2$, this contradicts the Nondeterministic Time Hierarchy Theorem.
-In practice, we use (integer) linear programming to find the best order of application
-of the rules, and the best parameters to use.
-
-We proceed similarly with properties of \ParSAT{} and \SharpSAT{}
-to prove results for fomulas with XOR gates.
+wlog, \ssym{} circuit only contains so-called ``Exact-Majority'' gates.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\section{Proof of thm}
-Speedup Rule for Alternating formulas.
-
-Slowdown Rule from algorithms for SAT assumptions:
-\begin{lemma}
-	Let $\Ff$ be a class of formulas closed under composition that can simulate projections in constant depth.
-	If $\SAT{}\in \Ff[c \log n]$, then $\NTIME[n] \subseteq \Ff[c \log n]$.
-\end{lemma}
-\begin{proof}
-	This follows from the fact that $\SAT{}$ is \NP-complete under uniform poly-log time first-order projections.
-\end{proof}
-
-Can we say the same for $\ParSAT$ and $\SharpSAT$?
-
-How to get the contradiction given the slowdown rule for $c < 4\cos(\pi /7)$.
-
-\subsection{Improving Brent/Spira}
-\todo[inline]{Prove this}
-Useful starting point : Size-Depth Tradeoffs for Boolean Formulae (Bonet, Buss).
-
-Remark: we must focus on the case when the input formula is a \NAND{} formula.
-Otherwise, we prove a much stronger result.
-Maybe prove the stronger result with a $2+\eps$ factor? Not very useful though.
-If we can go below 2, we improve something.
-%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%
-\section{Proof of optimality}
-\todo[inline]{Prove this}
+\section{Proof of main result}
 

### 2022-06-10 with_mip/main.py

@@ -23,10 +23,10 @@ def enumerate_annotations(n):
 
 def print_anno(q, r, x=None):
     s = '\\subseteq ' \
-        + ' '.join(f'({"E" if i%2 == 0 else "A"} n^{q[i]:.5f})' for i in range(len(q))) \
-        + f' Nand-depth[{r} log n]'
+        + ' '.join(f'({"E" if i%2 == 0 else "A"} n^{q[i]:.3f})' for i in range(len(q))) \
+        + f' Nand-depth[{r:.3f} log n]'
     if x is not None:
-        s += f'   (param {x:.5f})'
+        s += f'   (param {x:.3f})'
     print(s)
 
 def run_proof(annotation, params, p0, c):
@@ -51,7 +51,7 @@ def run_proof(annotation, params, p0, c):
             x = params[i+1]
             r = c*max(r-x, x, q[-1], q[-2]) if len(q) > 1 else c*max(r-x, x, q[-1])
             q.pop()
-            print_anno(q, r, x)
+            print_anno(q, r, None)
         elif annotation[i] == '2':
             r = c*max(r, q[-1], q[-2]) if len(q) > 1 else c*max(r, q[-1])
             q.pop()
@@ -86,16 +86,29 @@ def time_search(size, f, print_proof=False, **kwargs):
         run_proof(best_proof, best_x, best_r, best_c)
 
 def good_annotation(k):
-    return '1'*(k-1) + '100'*k
+    if k == 0:
+        return '0'
+    return '1' + good_annotation(k-1) + '10'*k + '0'
 
 def try_good_proof():
-    for s in [10,20,30,40,50]:
+    for s in [3,5,10,15,30,50]:
         t = time()
         c,p,x,r = best_param_sparse(good_annotation(s))
         t2 = time()
         print(f'Best c found: {c:.5f} in {(t2-t):.5f}s for annotation {p}')
 
+
+def try_proof(p):
+    t = time()
+    c,p,x,r = best_param_sparse(p, verbose=1)
+    t2 = time()
+    print(f'Best c found: {c:.5f} in {(t2-t):.5f}s for annotation {p}')
+
+
 if __name__ == '__main__':
+    # try_good_proof()
+    # try_proof('11010100100')
+    # exit(0)
     parser = argparse.ArgumentParser()
     parser.add_argument("depth", help="Number of 'up' steps in target proofs", type=int)
     parser.add_argument('-v', '--verbose', help="Activate debug of search functions", action='count', default=0)

### 2022-06-10 with_mip/pysearch.py

@@ -56,6 +56,7 @@ def best_proof_sparse(k, c_lb=1.0, c_ub=5.0, tol=0.001, n_threads=10, verbose=0)
         #       in order for the proof to be "well-typed".
         #   Copy l-1 quantifiers
         m.addConstrs((sd[i] == 1) >> (a[i, j] == a[i-1, j-1]) for j in range(1, nc) for i in range(2, nl))
+        # m.addConstrs((sd[i] == 1) >> (x[i] == r[i-1]/2) for i in range(2, nl)) # Best SD param is runtime/2 ?
         #   Set runtime. We wrap this into a function.
         def set_c_constraints(c):
             cx = []

### 2022-06-10 sym_eval/main.tex
Symsym cont

@@
 
+\newcommand{\FF}{\mathbb{F}}%
 \newcommand{\NN}{\mathbb{N}}%
@@
 \begin{theorem}
-	For all $n, m$, \ssym{} circuits of size $m$ over $n$ variables can be evaluated
+	For all $n, m$, \ssym{} circuits of size $m$ over $n$ variables, with possibly negated inputs, can be evaluated
 	on every input in time $O^*(2^n + m)$.
@@
 
-\begin{theorem} (??)
-	With our rules, we cannot do better than $4 \cos(\pi/7)$.
-\end{theorem}
-
 %%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{definition}[Symmetric functions]
-	A boolean function $f: \bit^n \mapsto \bit$ is \textit{symmetric} if there exists a function
-	$g: \{0,\ldots, n\} \mapsto \bit$ such that for every $x\bit^n$:
-	\[f(x) = g(\sum_{i=1}^n x_i).\]
+	A boolean function $f: \bit^n \rightarrow \bit$ is \textit{symmetric} if there exists a function
+	$g: \{0,\ldots, n\} \rightarrow \bit$ such that for every $x\in\bit^n$:
+	\[f(x) = g\left(\sum_{i=1}^n x_i\right).\]
 \end{definition}
+\todo[inline]{maybe it's easier to write everything over -1,1 ?}
 
@@
 
+For a symmetric function $f$ and its associated function $g$, we define $T(f) = g^{-1}(1)$,
+that is, the set of integers $t$ such that $f$ is one when exactly $t$ inputs of $f$ are equal to 1.
+
+Additionnally, for a gate $f$, we define $S^+(f)$ (resp. $S^-(f)$)
+to be the set of variables that are positive (resp. negative) inputs to $f$.
+
+\todo[inline]{What if a variables goes TWICE into a gate ? Doesn't matter for CNFs, but otherwise this needs to be a multiset, and we would have $X_i^{2a}$}
+
 wlog, \ssym{} circuit only contains so-called ``Exact-Majority'' gates.
+I don't even need this though.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Encoding \ssym{} as polys}
+\[\sum_{a = 1}^q \sum_{t = -n}^n g^{-at} \sum_{f : t\in T(f)} (\prod_{i \in S^+(f)} X_i^a) (\prod_{i \in S^-(f)} X_i^{-a})\]
+
+1. show that $P(x) = $ number of bottom layer gates equal to 1.
+
+Evaluate over $\{g, g^{-1}\}^n$ in $\FF_p$, where $q\mid p-1$.
+How to find $p,q$ so that we don't blow up the running time.
+
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Fast multipoint evaluation of these polys.}
+Since $q, t = O(n)$, we can iterate over outer loops:
+\[ P = \sum_{a = 1}^q \sum_{t = -n}^n g^{-at} P_{a,t}(X)\]
+We actually compute $2P$, can then divide by 2.
+Fix $t$, we note $P_a$ for $P_{a,t}$.
+Notation $P_{-a}$ for $P_{q-a}$ (same since $g$ has order $q$).
+
+Notice that for every $x\in \{g, g^{-1}\}^n$, we have $P_{-a}(X) = P_{a}(X^-1)$.
+
+TODO : needs index i as well.
 %%%%%%%%%%%%%%%%%%%%%%%%%

### 2022-06-15 sym_eval/main.tex
Updated sym eval, but there is an issue

--- 2022-06-10 sym_eval/main.tex
@@
 \newcommand{\bit}{\{0,1\}}%
+\newcommand{\mbit}{\{-1,1\}}%
+\newcommand{\gbit}{\{g^{-1},g\}}%
 \DeclareMathOperator*{\poly}{poly}
@@
 As a consequence, we obtain a $O^*(2^n)$-time algorithm that computes the
-truthe table of $2^n$-size CNFs.
+truthe table of CNFs with $2^n$ clauses.
 Lower bounds ensue. (do they though?)
 \end{abstract}
+\todo[inline]{Todo: issue with weights?}
 
@@
 
-This yields a significant improvement over the exhaustive search algorithm, that takes time $O(2^nm)$.
+This yields a significant improvement over the exhaustive search algorithm, that takes time $O(2^n\cdot m)$.
 
 \begin{corollary}
-	For any $c < 4 \cos(\pi/7)$, \SAT{} does not have \NAND{} formulas of \textbf{size} $n^{c + o(1)}$.
+	CNFs (and DNFs) on $n$ variables and $m$ clauses can be evaluated
+	over all inputs in time $O^*(2^n + m)$.
 \end{corollary}
-
-\begin{theorem} (??)
-	With our rules, we cannot do better than $4 \cos(\pi/7)$.
-\end{theorem}
 
@@
 \begin{definition}[Symmetric functions]
-	A boolean function $f: \bit^n \mapsto \bit$ is \textit{symmetric} if there exists a function
-	$g: \{0,\ldots, n\} \mapsto \bit$ such that for every $x\bit^n$:
-	\[f(x) = g(\sum_{i=1}^n x_i).\]
+	A boolean function $f: \mbit^n \rightarrow \mbit$ is \textit{symmetric} 
+	if there exists a function
+	$h: \{-n,\ldots, n-1, n\} \rightarrow \mbit$ such that for every $x\in\mbit^n$:
+	\[f(x) = h\left(\sum_{i=1}^n x_i\right).\]
+	The function $h$ is called the activation function of $f$.
 \end{definition}
@@
 \begin{definition}[\ssym{} circuits]
+	A \ssym{} circuit is a depth 2 circuit where every gate computes a symmetric function.
+	Inputs to the circuit can be negated.
 \end{definition}
+Notice that we can assume w.l.o.g. that the output of the gates of the bottom layer is not negated, since the negation of a symmetric function is also a symmetric function.
 
-wlog, \ssym{} circuit only contains so-called ``Exact-Majority'' gates.
+For a symmetric function $f$ and its associated function $h$, we define $T(f) = h^{-1}(1)$,
+that is, $T(f)$ is the set of integers $t$ such that $f$ is 1 when exactly $t$ of its inputs are equal to 1.
+
+Additionnally, for a gate $f$, we define $S^+(f)$ (resp. $S^-(f)$)
+to be the set of variables that are positive (resp. negative) inputs to $f$.
+Finally, let $S(f) = S^+(f) \cup S^-(f)$.
+
+\todo[inline]{TODO: Fix. What if a variables goes TWICE into a gate ? Doesn't matter for CNFs, but otherwise this needs to be a multiset, and we would have $X_i^{2a}$}
 
@@
 \section{Proof of main result}
+Given a \ssym{} circuit $C$ and a enumeration $F$ of the gates on its second layer, we show how to construct a polynomial $P_C$ 
+whose value at $(g^{x_i})_{i =1,\ldots n}$ is equal to $\sum_{f \in F} f(x)$.
+By applying the activation function of the topmost gate of $C$ to the 
+value of $P_C$, we can recover the value of $C$.
+Here, $g$ is a well chosen element that generates a small multiplicative subgroup of $\FF_p$.
+
+We then show how to evaluate efficiently $P_C$ on all inputs of $\gbit^n$,
+from which we can compute the truth table of $C$.
+
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Encoding \ssym{} as polynomials}
+
+Let $C$ be a \ssym{} circuit, and let $F$ be an enumeration of the symmetric gates on its second layer. We encode it with the following polynomial:
+
+\[P_C(X) = -|F| + 2q^{-1}\sum_{a = 1}^q \sum_{t = -n}^n g^{-at} \sum_{f \in F : t\in T(f)} \left(\prod_{i \in S^+(f)} X_i^a\right) \left(\prod_{i \in S^-(f)} X_i^{-a}\right)\]
+
+We will evaluate $P_C$ over the finite field $\FF_p$, where $p$ is a prime number larger that $m$. 
+Moreover, $q$ is a prime number larger that $2n$ that divides $p-1$, 
+and $g$ is an element of order $q$ in $\FF_p$. 
+See Section~\ref{sec:primes} for a discussion of the existence of suitable values for $p$ and $q$.
+
+\begin{lemma}\label{lemma:symgate}
+	For any symmetric function $f$ and for any $x\in\mbit^n$, we have:
+	\[f(x) = -1 + 2q^{-1}\sum_{t \in T(f)} \sum_{a = 1}^q g^{-at} \cdot \prod_{i \in S(f)} g^{ax_i}\]
+\end{lemma}
+\begin{proof}
+	For a fixed $t$, let $P_t(x) = \sum_{a = 1}^q g^{-at} \cdot \prod_{i \in S(f)} g^{ax_i}$.
+	We show the following:
+	\[P_t(x) = \begin{cases}
+		q & \text{ if } \sum_{i=1}^n x_i = t,\\
+		0 & \text{ otherwise.}
+	\end{cases}\]
+
+	First, notice that if $\sum_{i=1}^n x_i = t$, 
+	then $g^{-at} \cdot \prod_{i \in S(f)} g^{ax_i} = g^{-at} g^{a \sum_{i=1}^n x_i} = 1$, 
+	and therefore $P_t(x) = q$.
+
+	Now, if $\sum_{i=1}^n x_i \neq t$, then $g^{-at} \cdot \prod_{i \in S(f)} g^{ax_i} = g^{ac}$ for some $c \neq 0 \mod q$.
+	Therefore, we have:
+	\begin{flalign*}
+		P_t(x) 	&= \sum_{a=1}^q g^{ac}\\
+				&= g^c\cdot(g^{qc} - 1)\cdot(g^c - 1)^{-1}\\
+				&= 0 \text{ since $g$ has order $q$.}
+	\end{flalign*}
+
+	Therefore, $\sum_{t \in T(f)} \sum_{a = 1}^q g^{-at} \cdot \prod_{i \in S(f)} g^{ax_i} = \sum_{t \in T(f)} P_t$ is equal to $q$ if $\sum_{i=1}^n x_i \in T(f)$, and 0 otherwise.
+	Multiplying by $2q^{-1}$ and substracting 1 yields the value of $f(x)$.
+\end{proof}
+
+\begin{lemma}
+	Let \[P_C(X) = -|F| + 2q^{-1}\sum_{a = 1}^q \sum_{t = -n}^n g^{-at} \sum_{f \in F : t\in T(f)} \left(\prod_{i \in S^+(f)} X_i^a\right) \left(\prod_{i \in S^-(f)} X_i^{-a}\right)\]
+	Then, for every $x\in\mbit^n$, we have:
+	\[P_C(g^{x_1}, \ldots, g^{x_n}) = \sum_{f \in F} f(x).\]
+\end{lemma}
+\begin{proof}
+	Use Lemma~\ref{lemma:symgate}, plug in $X_i = g^{x_i}$,
+	take into account the fact that some inputs may be negated, 
+	sum over $f\in F$, and reverse summation order.
+\end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Moderate-size fields with small subgroups}
+Evaluate over $\{g, g^{-1}\}^n$ in $\FF_p$, where $q\mid p-1$.
+How to find $p,q$ so that we don't blow up the running time.
+
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Fast multipoint evaluation of these symmetric polynomials}
+In what follows, let $P_{a, t}$ denote the polynomial
+\[P_{a, t}(X) = g^{-at}\sum_{f \in F : t\in T(f)} \left(\prod_{i \in S^+(f)} X_i^a\right) \left(\prod_{i \in S^-(f)} X_i^{-a}\right).\]
+We use $P_{-a, t}$ to denote $P_{q-a, t}$: this is barely an abuse of notation
+since all of our computation takes place with $X\in\gbit^n$ 
+and $g$ has order $q$.
+
+Notice that for every $X\in\gbit^n$, we have $P_{-a, t}(X) = g^{2at}P_{a, t}(X^-1)$,
+where $X^{-1} = (X_i^{-1}; i = 1, \ldots, n)$. 
+Therefore, for every $X\in\gbit^n$, we now rewrite $P_C$ as:
+\begin{flalign*}
+	P_C(X) 
+	&= -|F| + 2q^{-1}\sum_{t = -n}^n \sum_{a=1}^q P_{a, t}(X)\\ 
+	&= -|F| + q^{-1}\sum_{t = -n}^n \left(\sum_{a=1}^q P_{a, t}(X) + \sum_{a=0}^{q-1} P_{a, t}(X)\right)\\ 
+	&= -|F| + q^{-1}\sum_{t = -n}^n \left(\sum_{a=1}^q P_{a, t}(X) + \sum_{a=1}^{q} P_{-a, t}(X)\right)\\ 
+	&= -|F| + q^{-1}\sum_{t = -n}^n \sum_{a=1}^q P_{a, t}(X) + P_{-a, t}(X)\\
+	&= -|F| + q^{-1}\sum_{t = -n}^n \sum_{a=1}^q P_{a, t}(X) + P_{-a, t}(X)
+\end{flalign*}
+To compute $P_C$ over all inputs in $\gbit^n$, it suffices to compute the values of $P_{a, t}$ over all inputs for every $a,t$. 
+This only adds polynomial overhead since there are at most $2n+1$ values of $t$ and $q = O(n)$.
+
+In what follows, fix a value of $t$, and let $P_a$ denote the polynomial $P_{a,t}$.
+We give a recursive algorithm to evaluate $P_a + P_{-a}$ over all inputs.
+At the $i$-th level of the recursion, we branch on the variable $X_i$.
+The polynomial $P_a$ can be decomposed as a sum as follows:
+\[P_a(X) = Q_0 + X_i^a Q_{a,+} + X_i^{-a} Q_{a,-}\]
+Similarly, for $P_{-a}$:
+\[P_{-a}(X) = Q_0 + X_i^{-a} Q_{-a,+} + X_i^{a} Q_{-a,-}\]
+Here, again, we have $Q_{-a, +}(X) = g^{2a} Q_{a, +}(X^{-1})$, 
+and $Q_{a, -}(X) = g^{-2a} Q_{-a, -}(X^{-1})$.
+Therefore, we have:
+\begin{flalign*}
+	P_a(X) 		&= Q_0(X) + X_i^a Q_{a,+}(X) + X_i^{-a} Q_{a,-}(X^{-1})\\
+	\text{and } 
+	P_{-a}(X) 	&= Q_0(X) + X_i^{-a} g^{2a} Q_{a,+}(X^{-1}) + X_i^{a} g^{2a} Q_{a,-}(X^{-1})\\
+\end{flalign*}
+%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%
+% \section{Evaluating weighted \ssym{} circuits}
+% One way to deal with weighted circuits might be the following:
 

### 2022-06-15 article/main.tex
Almost done ?

@@
 \newcommand{\Poly}{\textsf{P}}%
+\newcommand{\Log}{\textsf{L}}%
 \newcommand{\NP}{\textsf{NP}}%
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\title{Uniform formula lower bounds for finding and counting satisfiable assignments}
+\title{Uniform formula lower bounds for testing the existence of and counting satisfiable assignments}
 \author{Gabriel Bathie, Ryan Williams}
@@
 \end{abstract}
-
-List of things to do:
-\begin{itemize}
-	\item Improve Spira/Brent to show that a $c \log n$ depth 
-	lower bound implies an $n^{c-\eps}$ size lower bound?
-	\item Show that this bound is optimal w.r.t. these rules?
-\end{itemize}
 
@@
 
-Related work:
-- Original line of work : \cite{fortnow2000time,fortnow2005time}
-- Main line of work \cite{williams2006inductive,williams2007time,williams2013alternation} and its dramatic resolution \cite{buss2015limits}.
-- Broader classes \cite{mudigonda2020time}
+A line of work initiated by Fortnow~\cite{fortnow2000time,fortnow2005time} in 1997 
+has proven model-independent time-space lower bounds for the SAT
+problem and related problems within the polynomial-time hierarchy.
+This approach was then extended by Williams~\cite{williams2006inductive,williams2007time,williams2013alternation} 
+into a complex framework which culminated in proving that \SAT{} 
+does not have a small-space algorithm that runs in time $n^{2\cos(\pi/7)} \simeq n^{1.801}$.
+This line of work concluded~\cite{buss2015limits} with a proof 
+that the $2\cos(\pi/7)$ exponent is optimal for this framework,
+which implies that radically new ideas would be required to separate \NP{} from \Log{}.
+This approach was later expanded to obtain slightly stronger lower bounds
+for quantum classes, and against small-space randomized computation~\cite{mudigonda2020time}.
 
-Best known formula lower bounds for languages (functions?) in NP: 
-\begin{itemize}
-	\item $\Omegat{n^3}$ (for both uniform and non-uniform formulas) for Andreev's function \cite{hastad1998shrinkage}\todo{check source} in the De Morgan basis $\{\wedge, \vee, \neg\}$.
-	\item $\Omega(n^2/\log n)$ in the full binary basis (Nechiporuk)
-	\item $n^{\Theta(\log n)}$ for monotone formulas (formulas over the $\{\wedge, \vee\}$ basis).
-\end{itemize}
+In this work, we explore a new extension of this line of work, 
+which gives lower bounds on the size uniform formulas that solve \SAT{}
+and related problems, such as \ParSAT{} and \SharpSAT{}.
+By giving an efficient simulation of uniform log-depth \NAND{} formulas 
+by $\Sigma_2$ computation, we are able to show that \SAT{}
+does not have \textit{uniform} \NAND{} formulas 
+of depth less than~$4 \cos(\pi/7) \log n \simeq 3.603 \log n$.
 
+TODO: We also extend this result to show that \SharpSAT{} does not have
+alternating AND-OR/XOR-EQ formulas of depth less than~$3.603 \log n$.
+\todo[inline]{does that extend to $n^c$-size lower bound for some $c > 3$?}
+
+Finding strong lower bounds on the formula size of explicit functions in \NP{}
+is a long standing open problem.
+The best lower bounds known depend on the computational basis, that is, 
+the set of gates that are allowed in the formula.
+For the full binary basis $B_2$, Nechiporuk\cite{nechiporuk66boolean}
+proved in 1966 an $\Omegat{n^2}$ lower bound for an explicit function in \NP{}, 
+which remains unbeaten to this day.
+For the De Morgan basis $\{\wedge, \vee, \neg\}$, Andreev~\cite{andreev1987method} 
+proved an $\Omegat{n^{2.5}}$ lower bound, 
+which was later improved by H{\aa}stad~\cite{hastad1998shrinkage} to $\Omegat{n^{3}}$.
+This lower bound is optimal up to polylog factors,
+since their function has (uniform) formulas of size $\Ot{n^{3}}$.
+% Another interesting case is that of mononotone formulas, i.e. formulas over the $\{\wedge, \vee\}$ basis,
+% for which Razborov, Alon and Boppana~\cite{razborov1985lower, alon1987monotone} showed
+% a superpolynomial $n^{\Theta(\log n)}$ lower bounds for the \NP-complete \textsf{Clique} problem.
+\todo{add monotone circuits lower bounds? But that's not formulas}
+
+Note that all the lower bounds mentioned in the previous paragraph 
+are \textit{size} lower bounds of the form $n^c$,
+and they hold for \textit{non-uniform} formulas.  
+The lower bounds that we give are \textit{depth} lower bounds of the form $c\log n$,
+and only hold for \textit{uniform} formulas.
+While a $n^c$ size lower bound implies a $c\log n$ depth lower bounds,
+the converse is not immediately true; 
+one may get an $n^c$ size lower bound from an
+$\alpha c\log n$ depth lower bound, where $\alpha$ depends on the computation basis,
+using the work of Brent~\cite{brent1974parallel} and Spira~\cite{spira1971time}.
+\todo{can we show something here? e.g. improve for $B_2$?}
+
+%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Results overview}
@@
 
+Namely, we prove the following:
 \begin{theorem}
 	For any $c < 4 \cos(\pi/7) \simeq 3.603$, \SAT{} does not have \NAND{} formulas of depth $(c + o(1)) \log n$.
-\end{theorem}
-
-\todo[inline]{Prove all of the following.
-
-After checking \cite{sergeev2019relation}, this is probably incorrect...}
-\begin{lemma}
-	For any \NAND{} formula on $n$ inputs of size $n^c$ and any $\eps > 0$, 
-	there exists an equivalent \NAND{} formula of depth $(c+\eps) \log n$.
-\end{lemma}
-
-\begin{corollary}
-	For any $c < 4 \cos(\pi/7)$, \SAT{} does not have \NAND{} formulas of \textbf{size} $n^{c + o(1)}$.
-\end{corollary}
-
-\begin{theorem} (??)
-	With our rules, we cannot do better than $4 \cos(\pi/7)$.
 \end{theorem}
@@
 	if there exists a \textbf{RAM} machine $B_\varphi$
-	such that for every $n$, every $x\in\bit^n$ and every $b\in\bit^{\leq (c + o(1))\log n}$,
+	such that for every $n$, every $x\in\bit^n$ and every 
+	$b\in\bit^{\leq (c + o(1))\log n}$,
 	$B_\varphi(x, b, n)$ outputs a description of the gate at position $b$ in $\varphi_n$ in time $\log^{O(1)} n$.
@@
 class $(Q_1~n^{a_1})\ldots(Q_k~n^{a_k})~\Cc$, 
-where $Q_i$ is one of $\forall, \exists$ or $\oplus$.
+where $Q_i$ is one of $\forall$ or $\exists$
+ % or $\oplus$.
 A computation in a quantified complexity class is defined as follows:
@@
 		\item $Q_i$ is $\forall$, and for every $y_i$, the computation
-		accepts $x_{i}$ starting from stage $i+1$, 
-		\item $Q_i$ is $\bigoplus$, the number of $y_i$ such that the computation
-		accepts $x_{i}$ starting from stage $i+1$ is odd. 
+		accepts $x_{i}$ starting from stage $i+1$. 
+		% \item $Q_i$ is $\bigoplus$, the number of $y_i$ such that the computation
+		% accepts $x_{i}$ starting from stage $i+1$ is odd. 
 	\end{itemize}
@@
 allows us to deduce that the same is holds for all of $\NTIME[n]$, 
-a result referred to as the ``Slowdown rule''.
-\begin{lemma}[Slowdown rule]
-	% Let $\Ff$ be an expressive\todo{define} class of formulas, and let $c > 1$.
+\begin{lemma}\label{lemma:usual_sd}
 	If \SAT{} is in $\NDL{c}$ for some $c>0$,
@@
 \end{lemma}
-\todo[inline]{That's not the rule...}
 
@@
 	For every $c > 0, x \leq c$, we have:
-	\[\textsf{NAND-depth}[c \log n] \subseteq (\exists n^{x/2}) (\forall n^1) \textsf{NAND-depth}[(c-x) \log n]\]
+	\[\NDL{c} \subseteq (\exists n^{x/2}) (\forall n^1) \NDL{(c-x)}\]
 \end{lemma}
 
-We then combine multiple application of these lemmas with various parameters,
-with the goal of contradicting a known hierarchy theorem.
-Assuming that $\NTIME[n] \subseteq \NDL[c]$ for some $n$, we obtain the following hierarchy theorem
-for NAND formulas, as a corollary of the Nondeterministic Time Hierarchy Theorem:
+The above lemma can be combined with exhaustive search and Lemma~\ref{lemma:usual_sd} 
+to yield an improved version of Lemma~\ref{lemma:usual_sd} for quantified classes, 
+which we call the ``Slowdown rule''.
+\begin{lemma}[Slowdown rule]
+	Assume that \SAT{} is in $\NDL{c}$ for some $c>0$.
+	Then, for every $a,b,d > 0$, we have:
+	\[\ldots (Q~n^{a}) (\neg Q~n^b) \NDL{d}
+	\subseteq \ldots (Q~n^{a}) \NDL{c\cdot\max(d/2, a, b)}.\]
+	% \[\ldots (Q_{k-1}~n^{a_{k-1}}) (Q_k~n^{a_k}) \NDL{d} 
+	% \subseteq (Q_{k-1}~n^{a_{k-1}}) \NDL{c\cdot \max(d/2, a_k, a_{k-1})}\]	
+\end{lemma}
+
+To prove our results, 
+we combine multiple applications of these rules with suitable parameters,
+with the goal of contradicting a hierarchy theorem.
+In this work, we aim to reach a contradiction of the following hierarchy
+theorem for NAND formulas, that we obtain from the assumption that
+$\NTIME[n] \subseteq \NDL[c]$ for some $c$, 
+as a corollary of the Nondeterministic Time Hierarchy Theorem:
 \begin{theorem}\label{thm:nandh}
@@
 \end{flalign*}
-For $c^2/2  < 4 \leftrightarrow c < 2\sqrt{2} \simeq 2.8284\ldots$, this contradicts Theorem~\ref{thm:nandh}.
+For $c^2/2  < 4 \Leftrightarrow c < 2\sqrt{2} \simeq 2.8284\ldots$, this contradicts Theorem~\ref{thm:nandh}.
 
-In practice, we use (integer) linear programming to find the best order of application
-of the rules, and the best parameters to use.
+In practice, we use (integer) linear programming to find the
+best sequence of rules to apply, along with the optimal parameters.
+See Appendix~\ref{app:ilp} for a detailed presentation of the models that we use.
 
@@
 \end{lemma}
+\todo[inline]{fix quantifiers?}
 
-The intuitive idea behind this lemma is that the only information needed to know that an OR gate evaluates
+The intuitive idea behind this lemma is that the only
+information needed to know that an OR gate evaluates
 to 1 is that one of its children evaluates to 1. 
-We apply this idea to every OR gate in the first $x \log n$ levels of the formula,
+We apply this idea to every OR gate in the first $x \log n$
+levels of the formula,
 and then use a universal quantifier to handle the remaining AND gates.
 
-The role of the existential quantifier is to guess for every OR gate in the first $(x/2)\log n$ levels,
+The role of the existential quantifier is to guess for every OR gate in the first $x\log n$ levels,
 which of its children evaluates to 1, and we only evaluate the corresponding subformula.
 Moreover, making this guess reduces the number of OR gates that we will evaluate
-(this number doubles every two levels, instead of doubling at every level), 
+(since the formula alternate between levels of OR and AND gates, this number doubles every two levels, instead of doubling at every level), 
 therefore we only need to guess $O(n^{x/2})$ bits.
@@
 We are left with a formula starting with $(x/2)\log n$ levels of AND gates, 
-followed by OR/AND formulas of depth $(d-x + o(1)) \log n$.
+followed by smaller OR/AND formulas of depth $(d-x + o(1)) \log n$.
 The main formula evaluate to 1 if and only if for every of these $O(n^{x/2})$
@@
 	We prove the result for $\exists\forall$ computations, 
-	the result for $\forall\exists$ computations follows from the fact that $\NDL{d}$ is closed
-	under complementation.
+	the result for $\forall\exists$ computations follows from the fact
+	that $\NDL{d}$ is closed under complementation.
 
-	Let $L\in\NDL{d}$, let be $M : (x,i,n) \mapsto g$ be the a descriptor machine for $L$,
-	and let $x$ be an input of length $n$.
+	Let $L\in\NDL{d}$, let $M : (x,i,n) \mapsto g$ be 
+	a descriptor machine for $L$, and let $x$ be an input of length $n$.
 
@@
 	Now, using an exhaustive search over the $n^{d/2}$ bit strings of length $\frac{d}{2}\log n$
-	followed by a deterministic $n^{o(1)}$ computation,
-	we get
-	$\NDL{d} \subseteq (\exists n^{d/2}) \TIME[n^{d/2}]$.
+	followed by a deterministic $n^{o(1)}$ computation, we get that
+	$\NDL{d} \subseteq (\exists n^{d/2}) \TIME[n^{d/2}] = \NTIME[n^{d/2}]$.
+
+	The inclusion in $\coNTIME[n^{d/2}]$ follows from the stability of $\NDL{d}$
+	under complement.
 \end{proof}
 
-We now discuss
+We now discuss how to efficiently remove quantifiers from an alternating \ND{} computation.
 Slowdown rule from algorithms for SAT assumptions:
@@
 Since the previous result only holds for $\NTIME[n^d], d \ge 1$, 
-we need to take into
-account the edge case where the running time of the $i+1$-th
+we need to take into account the edge case where the running time of the $i+1$-th
 stage is sublinear in the size of its input.
+
+We will now show that applying the Speedup rule once and
+using exhaustive search to remove the logarithmic-size quantifier
+yields an improved Slowdown rule for quantified classes.
 We have the following:
-\begin{corollary}[Extended Slowdown rule]
+\begin{corollary}[Extended Slowdown rule]\label{cor:extended_slowdown}
 	If $\SAT{}\in \NDL{c}$, then for every $d > 1$, we have:
@@
 	\subseteq \ldots (Q~n^{a}) \NDL{c\cdot\max(d/2, a, b)}.\]
-	\todo{fix the $d/2$}
 \end{corollary}
@@
 	Without loss of generality, assume that $Q = \forall$, and therefore $\neg Q = \exists$.
-	In that case, we have $(\neg Q~n^b) \NDL{d} \subseteq \NTIME[n^{\max(d, b)}]$,
-	since evaluating a formula of depth $d \log n$ can be done in deterministic time
-	$n^{d + o(1)}$.
+	In that case, for all $q, r > 0$, we have $(\neg Q~n^q) \NTIME[n^r] \subseteq \NTIME[n^{\max(q,r)}] $.
+	Applying Corollary~\ref{cor:speedup} on the $\NDL{d}$ part and then using the 
+	above observation leads to the following:
+	\begin{flalign*}
+	\ldots (Q~n^{a}) (\neg Q~n^b) \NDL{d} 
+	&\subseteq \ldots (Q~n^{a}) (\neg Q~n^b) \NTIME[n^{d/2}]\\	
+	&\subseteq \ldots (Q~n^{a}) \NTIME[n^{\max(d/2, d)}]	
+	\end{flalign*}
 
-	Now, consider the $\NTIME[n^{\max(d, b)}]$ language $L$ that correspond to the last stage of the alternating 
+	Now, consider the $\NTIME[n^{\max(d/2, b)}]$ language $L$
+	that correspond to the last stage of the alternating 
 	computation, and its corresponding nondeterministic machine $M$. 
 	It takes as input a string of length $m = n^{a}$, and its runtime, as a function of $m$, 
-	is $m^{\max(d, b)/a}$.
-	We apply the Slowdown rule to $M$. There are two possible cases:
+	is $m^{\max(d/2, b)/a}$.
+	We apply the Slowdown rule (Lemma~\ref{lemma:slowdown}) to $M$. There are two possible cases:
 	\begin{itemize}
-		\item if $\max(d, b)/a < 1$, then we have $L \in \NTIME[m^{\max(d, b)/a}] \subseteq \NTIME[m]
+		\item if $\max(d/2, b)/a < 1$, then we have 
+		$L \in \NTIME[m^{\max(d/2, b)/a}] \subseteq \NTIME[m]
 		\subseteq \ND[c \log m] \subseteq \NDL{c \cdot a}$.
-		Since  $\max(d, b) < a$, we have $a = \max(d, a, b)$ and therefore 
-		$\NDL{c \cdot a} = \NDL{c\cdot\max(d, a, b)}$.
-		\item otherwise, $\max(d, b) \ge a$, and applying the slowdown rule yields
-		$L \in \NTIME[m^{\max(d, b)/a}] =  \NTIME[n^{\max(d, b)}] \subseteq \NDL{c \cdot \max(d, a, b)}$.
+		Since  $\max(d/2, b) < a$, we have $a = \max(d/2, a, b)$ and therefore 
+		$\NDL{c \cdot a} = \NDL{c\cdot\max(d/2, a, b)}$.
+		\item otherwise, $\max(d/2, b) \ge a$, and applying the slowdown rule yields
+		$L \in \NTIME[m^{\max(d/2, b)/a}] =  \NTIME[n^{\max(d/2, b)}] \subseteq \NDL{c \cdot \max(d/2, a, b)}$.
 	\end{itemize}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Structure of optimal proofs}
 How to get the contradiction given the slowdown rule for $c < 4\cos(\pi /7)$.
 
-\subsection{Improving Brent/Spira}
+%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%
+\section{Improving Brent/Spira}
 \todo[inline]{Prove this}
@@
 If we can go below 2, we improve something.
+
 %%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%
-\section{Proof of optimality}
-\todo[inline]{Prove this}
+%%%%%%%%%%%%%%%%%%%%%%
+\appendix
+%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%
+\section{Program to find optimal parameters}\label{app:ilp}
+Improves over Williams (cite): no need to enumerate proof annotations.
+
+Therein, Williams enumerates all possible proof skeletons of length $k$,
+and tries to find to find the best set of parameters for each of them, using a linear program.
+In this work, we propose a different approach, that proves to be more efficient in practice:
+we use an Integer Linear program whose solution is
+an optimal proof of length $k$
+along with its parameters.
+
+More precisely, we give an ILP $P_{c,k}$ which is feasible if and only if
+there exists a proof annotation $p$ of length $k$\todo{it's not exactly that, it's $k$ up steps} along with parameters $x, r_0$
+such that $p$ proves a contradiction under the assumption that $\SAT\in\NDL{c}$.
+Moreover, its solution encodes $p,q,r_0$.
+By using binary search over $c$, we can quickly get an approximation 
+of the largest $c$ such that a proof of length $k$ can reach a contradiction under $H_c$\todo{define "hypothesis c"}, together with a proof for $c-\eps$.
+Repeating the above steps for increasing values of $k$ shows that the sequence $c_k^*$\todo{define ?}
+Table of $c_k^*$ for first few values of $k$?
+Seems to show convergence for some value $c^\infty > 3.6$.
+
+\todohere describe model + link to code?
 

### 2022-06-24 article/main.tex
intro almost done

@@
 \newcommand{\ND}{\textsf{NDepth}}%
-\newcommand{\NDL}[1]{\ND[ #1 \log n]}%
+\newcommand{\NDL}[1]{\ND\left[ #1 \log n\right]}%
+\newcommand{\TS}[1]{\textsf{TS}\left[ n^{ #1 }\right]}%
 
 \newcommand{\todohere}{\todo[inline]{TODO}}%
+\newcommand{\todoh}[1]{\todo[inline]{TODO: #1 }}%
 
@@
 
-Motivation: understanding complexity. Explicit lower bounds : programs, circuits, formulas.
-Uniformity.
+\todoh{Motivation: understanding complexity. Explicit lower bounds: programs, circuits, formulas.
+Uniformity.}
 
@@
 which implies that radically new ideas would be required to separate \NP{} from \Log{}.
-This approach was later expanded to obtain slightly stronger lower bounds
+This approach was later extended to obtain slightly stronger lower bounds
 for quantum classes, and against small-space randomized computation~\cite{mudigonda2020time}.
@@
 In this work, we explore a new extension of this line of work, 
-which gives lower bounds on the size uniform formulas that solve \SAT{}
-and related problems, such as \ParSAT{} and \SharpSAT{}.
-By giving an efficient simulation of uniform log-depth \NAND{} formulas 
+which gives lower bounds on the depth of uniform formulas that solve \SAT{}.
+By giving a new efficient simulation of uniform log-depth \NAND{} formulas 
 by $\Sigma_2$ computation, we are able to show that \SAT{}
@@
 
-TODO: We also extend this result to show that \SharpSAT{} does not have
-alternating AND-OR/XOR-EQ formulas of depth less than~$3.603 \log n$.
-\todo[inline]{does that extend to $n^c$-size lower bound for some $c > 3$?}
-
 Finding strong lower bounds on the formula size of explicit functions in \NP{}
-is a long standing open problem.
+is a long-standing open problem.
 The best lower bounds known depend on the computational basis, that is, 
@@
 since their function has (uniform) formulas of size $\Ot{n^{3}}$.
-% Another interesting case is that of mononotone formulas, i.e. formulas over the $\{\wedge, \vee\}$ basis,
-% for which Razborov, Alon and Boppana~\cite{razborov1985lower, alon1987monotone} showed
-% a superpolynomial $n^{\Theta(\log n)}$ lower bounds for the \NP-complete \textsf{Clique} problem.
-\todo{add monotone circuits lower bounds? But that's not formulas}
 
 Note that all the lower bounds mentioned in the previous paragraph 
-are \textit{size} lower bounds of the form $n^c$,
+are \textit{size} lower bounds of the form $n^{c + o(1)}$
 and they hold for \textit{non-uniform} formulas.  
-The lower bounds that we give are \textit{depth} lower bounds of the form $c\log n$,
+The lower bounds that we give are \textit{depth} lower bounds of the form $(c + o(1))\log n$,
 and only hold for \textit{uniform} formulas.
-While a $n^c$ size lower bound implies a $c\log n$ depth lower bounds,
+While an $n^c$ size lower bound implies a $c\log n$ depth lower bounds,
 the converse is not immediately true; 
@@
 using the work of Brent~\cite{brent1974parallel} and Spira~\cite{spira1971time}.
-\todo{can we show something here? e.g. improve for $B_2$?}
 
@@
 
-We first study \NAND{} formulas. 
 The \NAND{} basis is rather expressive: Andreev's function has \NAND{} formulas of depth $(3+o(1)) \log n$ (and therefore of size $\Ot{n^3}$), which matches the lower bound for the De Morgan Basis.
+\todoh{more about nand?}
 
 Namely, we prove the following:
-\begin{theorem}
-	For any $c < 4 \cos(\pi/7) \simeq 3.603$, \SAT{} does not have \NAND{} formulas of depth $(c + o(1)) \log n$.
+\begin{theorem}\label{thm:main}
+	For any $c < 4 \cos(\pi/7) \simeq 3.603$, \SAT{} does not have 
+	uniform \NAND{} formulas of depth $(c + o(1)) \log n$.
 \end{theorem}
@@
 \subsection{Uniform formulas}
-A family of formulas
+A boolean formula $\phi$ over $n$ inputs is a 
+rooted full binary tree\footnote{\textit{Full} means that every node has either two or zero children.}
+whose internal nodes, called gates, are labeled by a function $f: \bit^2 \rightarrow \bit$,
+and leaves are labeled with a variable, $x_i$ its negation $\neg x_i$ for some $i, 1\le i \le n$,
+or a constant $c\in\bit$.
+The depth of a formula is the depth of the underlying tree.
 
-In this work, we focus on formulas with depth $d(n) = (c + o(1))\log n$, for some $c > 1$.
+In this work, we focus on \NAND{} formulas, that is formulas whose gates are all labeled 
+with the function $\NAND: x,y \mapsto \neg(x\wedge y)$.
+The De Morgan law implies that $\NAND(x, y) = \neg x \wedge \neg y$, 
+therefore \NAND{} formulas are equivalent to ``alternating OR/AND'' formulas (up to negating some inputs), 
+that is formulas whose gate with even-depth \footnote{The root has depth 0.} 
+are OR gates, and odd-depth gates are AND gates.
+This alternative alternating point of view will be useful 
+when describing a procedure to compute the value of a formula with alternating computation.
 
+\todoh{To introduce: family of formulas computing a language}
+% In this work, we focus on formulas with depth $d(n) = (c + o(1))\log n$, for some $c > 1$.
+However, there exists formulas of \textit{constant} depth that compute undecidable languages.
+In this work, we restrict our attention to uniform formulas, 
+that is family of formulas for which the description of the
+$n$-input formula can be computed efficiently.
+We first recall a usual definition of uniform formulas.
 \begin{definition}[``Usual'' polylog-time uniform formulas]
@@
 
-Here, the position of a gate is given by a bitstring of length at most $(c + o(1))\log n$ 
+Here, the position of a gate is given by a bit string of length at most $(c + o(1))\log n$ 
 that describes the path to follow to reach that gate from the root, 
-with $0$ meaning ``go to left child'' and $1$ ``go to right child''.
+with $0$ meaning ``go to the left child'' and $1$ ``go to the right child''.
 
-In this work, we consider a stronger model of uniform formulas, 
+However, in out work, we need to use a stronger model of uniform formulas, 
 where the descriptor machine $A_\varphi$ has random access to the input $x$ 
-at which we wish to evaluate the formula, but is still limited to polylogarithmic time.
-\begin{definition}[Input-aware polylog-time uniform formulas]\todo{is this a good name?}\label{def:unif}
+at which we wish to evaluate the formula but is still limited to polylogarithmic time.
+\begin{definition}[Input-aware polylog-time uniform formulas]\label{def:unif}
 	A family of formulas $(\varphi_n)_{n \ge 1}$ of depth $(c + o(1))\log n$ is \textit{polylog-time uniform}
@@
 but does not use its random access to the input, 
-therefore the lower bounds that we give against model also hold for the usual one.
+therefore the lower bounds that we give against this model also hold for the usual one.
 
-Moreover, this definition allows for an efficient simulation of uniform formulas by RAM machines, 
-due to the negligible overhead.
-Indeed, for every $c > 1$, we have $\ND[c \log n] \subseteq TS[n^c]$.
-
-We will see the need for this stronger definition in \todo{ref}, 
+We will see the need for this stronger definition in Section~\ref{sec:rules}, 
 where we need to give $O(\log n)$ bits of information to the descriptor
@@
 
-Uniform \NAND{}/alternating formulas. Specific model, comparison with the usual model.
+Finally, for any $c \ge 0$, we define $\NDL{c}$
+to be the set of languages that have uniform 
+input-aware polylog-time NAND formulas of depth$(c+o(1)) \log n$.
+
+Notice that uniform log-depth NAND formulas
+can be efficiently simulated by RAM machines, 
+due to the negligible runtime overhead of the descriptor.
+Indeed, for every $c > 1$, we have $\NDL{c} \subseteq TS[n^c]$.
 
@@
 
-We now introduce ``quantified'' complexity classes, which extend alternating classes 
-by adding a ``parity'' quantifier.
+We now introduce alternating complexity classes, a fine-grained version of the polynomial hierarchy.
 
 Given a complexity class $\Cc$ and constants $k, (a_i)_{i\leq k}$, 
-we consider \textit{quantified} complexity
+we consider \textit{alternating} complexity
 class $(Q_1~n^{a_1})\ldots(Q_k~n^{a_k})~\Cc$, 
 where $Q_i$ is one of $\forall$ or $\exists$
- % or $\oplus$.
-A computation in a quantified complexity class is defined as follows:
-\begin{definition}[Quantified computation]
+A computation in a alternating complexity class is defined as follows:
+\begin{definition}[Alternating computation]
 	A language $L$ is in $(Q_1~n^{a_1})\ldots(Q_k~n^{a_k})~\Cc$ 
@@
 		accepts $x_{i}$ starting from stage $i+1$. 
-		% \item $Q_i$ is $\bigoplus$, the number of $y_i$ such that the computation
-		% accepts $x_{i}$ starting from stage $i+1$ is odd. 
 	\end{itemize}
@@
 \end{definition}
-\todo[inline]{allow $O(\log n)$ quantifiers ?}
 
-\subsection{Overview of the methodology}
+\subsection{Overview of the methodology}\label{sec:overview}
 
 The general methodology used in this paper to obtain lower bounds
-extends the one used in~\cite{mudigonda2020time}.
+extends the one used in~\cite{williams2006inductive, mudigonda2020time}
+to the specific case of uniform \textsf{NAND} formulas of logarithmic depth.
 
-To prove that \SAT{} does not have formulas of depth less than $c\log n$, 
+To show that \SAT{} does not have formulas of depth less than $c\log n$, 
 we reason by contradiction.
 First, we assume that \SAT{} has such formulas, and the strong \NP-completeness of \SAT{}
-allows us to deduce that the same is holds for all of $\NTIME[n]$, 
+allows us to deduce that the same holds for all of $\NTIME[n]$, 
 \begin{lemma}\label{lemma:usual_sd}
@@
 
-We also show how to speed-up (or rather, reduce the depth of) computations by adding quantifiers, 
+We also show how to speed up (or rather, reduce the depth of) computations by adding quantifiers, 
 using various ``Speedup rules''. 
@@
 	\subseteq \ldots (Q~n^{a}) \NDL{c\cdot\max(d/2, a, b)}.\]
-	% \[\ldots (Q_{k-1}~n^{a_{k-1}}) (Q_k~n^{a_k}) \NDL{d} 
-	% \subseteq (Q_{k-1}~n^{a_{k-1}}) \NDL{c\cdot \max(d/2, a_k, a_{k-1})}\]	
 \end{lemma}
@@
 \end{flalign*}
-For $c^2/2  < 4 \Leftrightarrow c < 2\sqrt{2} \simeq 2.8284\ldots$, this contradicts Theorem~\ref{thm:nandh}.
+For $c^2/2  < 4 \Leftrightarrow c < 2\sqrt{2} \simeq 2.8284\ldots$,
+this contradicts Theorem~\ref{thm:nandh}.
 
@@
 
+\paragraph{Optimality of the result}
+In the case of \SAT{} vs small-space algorithms, 
+Buss and Williams~\cite{buss2015limits} proved that the result $\SAT\notin \TS{n^c}$
+for $c < 2\cos(\pi/7)$ is optimal for their framework. 
+It was later extended and simplified by Mudigonda and Williams~\cite{mudigonda2020time},
+who showed upper bounds on .
+It appears that their generalization covers our framework,
+therefore we cannot prove $\SAT\notin\NDL{c}$ for $c \ge 4\cos(\pi/7)$
+without finding a new (and improved) way to speed up \ND{} computation using alternations.
+
 %%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%
-\section{Proof of main result}
+\section{Proof of the main result}
 
-\subsection{Speedup and Slowdowns}
-We start by discussion the Speedup rule for Alternating formulas.
+\subsection{Speedup and Slowdowns}\label{sec:rules}
+We start by discussing the Speedup rule for Alternating formulas.
 
@@
 	\begin{flalign*}
-	\NDL{d} &\subseteq (\exists n^{x/2}) (\forall \frac{x}{2}\log n) \NDL{d-x}\text{ and}\\
-	\NDL{d} &\subseteq (\forall n^{x/2}) (\exists \frac{x}{2}\log n) \NDL{d-x}
+	\NDL{d} &\subseteq (\exists n^{x/2}) (\forall n) \NDL{d-x}\text{ and}\\
+	\NDL{d} &\subseteq (\forall n^{x/2}) (\exists n) \NDL{d-x}
 	\end{flalign*}
 \end{lemma}
-\todo[inline]{fix quantifiers?}
 
@@
 Moreover, making this guess reduces the number of OR gates that we will evaluate
-(since the formula alternate between levels of OR and AND gates, this number doubles every two levels, instead of doubling at every level), 
+(since the formula alternate between levels of OR and AND gates, this number doubles every two levels,
+instead of doubling at every level), 
 therefore we only need to guess $O(n^{x/2})$ bits.
@@
 followed by smaller OR/AND formulas of depth $(d-x + o(1)) \log n$.
-The main formula evaluate to 1 if and only if for every of these $O(n^{x/2})$
+The main formula evaluates to 1 if and only if for every of these $O(n^{x/2})$
 smaller formulas evaluates to true: 
-we can check this using a universal quantifiers, 
+we can check this using a universal quantifier, 
 which requires $O(\frac{x}{2}\log n)$ bits, and then running an $\NDL{d-x}$ computation.
+Note that while the second quantifier only requires $O(\frac{x}{2}\log n)$ nondeterministic bits,
+it still computes and outputs $n^{1+o(1)}$ bits, hence the $n^1$ exponent.
 
@@
 	the result for $\forall\exists$ computations follows from the fact
-	that $\NDL{d}$ is closed under complementation.
+	that $\NDL{d}$ is closed under taking the complement.
 
@@
 	We describe how to build bit strings $y$ of length $n^{x/2}$ and
-	\todohere 
+	\todoh{finish details of proof}
 \end{proof}
 
+Lemma~\ref{lemma:speedup} yields the following corollary,
+which gives us a simulation of formulas by Turing machines that is more efficient
+than the naive evaluation.
 \begin{corollary}\label{cor:speedup}
@@
 	Applying Lemma~\ref{lemma:speedup} with $x = d$, we obtain
-	$\NDL{d} \subseteq (\exists n^{d/2}) (\forall \frac{d}{2}\log n) \NDL{0}$.
-	Now, using an exhaustive search over the $n^{d/2}$ bit strings of length $\frac{d}{2}\log n$
-	followed by a deterministic $n^{o(1)}$ computation, we get that
+	$\NDL{d} \subseteq (\exists n^{d/2}) (\forall n) \NDL{0}$.
+	However, looking closely at the proof of Lemma~\ref{lemma:speedup}
+	reveals that the $\forall$ quantifier here only uses $O(\frac{d}{2}\log n)$
+	nondeterministic bits.
+	Therefore, we can use exhaustive search over the $n^{d/2}$ bit strings of length $\frac{d}{2}\log n$
+	to remove that quantifier.
+	This leaves us we a deterministic $n^{o(1)}$ computation, and we get that
 	$\NDL{d} \subseteq (\exists n^{d/2}) \TIME[n^{d/2}] = \NTIME[n^{d/2}]$.
@@
 
-We now discuss how to efficiently remove quantifiers from an alternating \ND{} computation.
-Slowdown rule from algorithms for SAT assumptions:
+We now discuss how to use this corollary to
+efficiently remove quantifiers from an alternating \ND{} computation.
+The first step is to show that the assumption $\SAT\in\NDL{c}$
+propagates efficiently to all of \NP{}.
+This result is known as the Slowdown rule: it allows us to remove a quantifier,
+at the cost of an increased depth of the verifier.
 \begin{lemma}[Slowdown rule]\label{lemma:slowdown}
@@
 with instances of size $n$ reduced to formulas of size $n^{1+o(1)}$.
-Intuitively, a first order projection is a depth 0 circuit, 
+Intuitively, a first-order projection is a depth 0 circuit, 
 i.e. each output bit is either a constant (0 or 1),
-an input $x_i$ or its negation, $\neg x_i$.
+an input $x_i$, or its negation $\neg x_i$.
 For a more in-depth exposition of first-order projections,
@@
 \begin{proof}[Proof of Lemma~\ref{lemma:slowdown}]
-	We first prove that $\SAT{}\in \NDL{c}$ implies that $\NTIME[n] \subseteq \NDL{c}$,
-	then use a padding argument to show that $\forall d\ge 1, \NTIME[n^d] \subseteq \NDL{c\cdot d}$,
-	and the result follows since $\NDL{c\cdot d}$ is closed under complement.
+	We first prove that $\SAT{}\in \NDL{c}$ implies that $\NTIME[n] \subseteq \NDL{c}$
+	and use a padding argument to show that 
+	\[\forall d\ge 1, \NTIME[n^d] \subseteq \NDL{c\cdot d}.\]
+	The result follows since $\NDL{c\cdot d}$ is closed under complement.
 
@@
 	uniform poly-log time first-order projections, 
-	\todo{ref for this? Folklore?}
 	with instances of size $n$ reduced to formulas of size $n^{1+o(1)}$.
-	Now, notice that the composition of a NAND formula of depth $d$ with a first order projection
+	Now, notice that the composition of a NAND formula of depth $d$ with a first-order projection
 	is also a NAND formula of depth $d$: it only consists in relabeling leaves of the formula.
@@
 	Therefore, we can use $(c+o(1)) \log n$-depth NAND formulas for \SAT{} on $n^{1+o(1)}$
-	inputs to decide $L$ by composing them with the first order projection; 
+	inputs to decide $L$ by composing them with the first-order projection; 
 	the resulting formula has size $(c+o(1)) \log n^{1+o(1)} = (c+o(1)) \log n$,
@@
 	calls $A$ on $x, b, n^d$.
-	If $A$ output a gate type or a constant, $B$ outputs the same information.
+	If $A$ outputs a gate type or a constant, $B$ outputs the same information.
 	If $A$ outputs a literal $x_i^*$, $B$ outputs $x_i^*$ if $i \le n$, and 1 otherwise.
@@
 Since the previous result only holds for $\NTIME[n^d], d \ge 1$, 
-we need to take into account the edge case where the running time of the $i+1$-th
-stage is sublinear in the size of its input.
+we need to take into account the edge case where the running time of
+the $i+1$-th stage is sublinear in the size of its input.
 
@@
 	Without loss of generality, assume that $Q = \forall$, and therefore $\neg Q = \exists$.
-	In that case, for all $q, r > 0$, we have $(\neg Q~n^q) \NTIME[n^r] \subseteq \NTIME[n^{\max(q,r)}] $.
+	In that case, for all $q, r > 0$, we have $(\exists n^q) \NTIME[n^r] \subseteq \NTIME[n^{\max(q,r)}] $.
 	Applying Corollary~\ref{cor:speedup} on the $\NDL{d}$ part and then using the 
@@
 	\begin{flalign*}
-	\ldots (Q~n^{a}) (\neg Q~n^b) \NDL{d} 
-	&\subseteq \ldots (Q~n^{a}) (\neg Q~n^b) \NTIME[n^{d/2}]\\	
-	&\subseteq \ldots (Q~n^{a}) \NTIME[n^{\max(d/2, d)}]	
+	\ldots (\forall n^{a}) (\exists n^b) \NDL{d} 
+		&\subseteq \ldots (\forall n^{a}) (\exists n^b) \NTIME[n^{d/2}]\\	
+		&\subseteq \ldots (\forall n^{a}) \NTIME[n^{\max(d/2, d)}]	
 	\end{flalign*}
@@
 %%%%%%%%%%%%%%%%%%%%%%
+\subsection{Conditional Depth hierarchy theorem for \textsf{NAND} formulas}
+
+We now prove the Conditional Depth hierarchy theorem, 
+stated without proof in Section~\ref{sec:overview}.
+\begin{theorem}
+	If there exists $c > 0$ such that $\SAT \subseteq \NDL{c}$, 
+	then for all $b > a > 0$, we have
+	\[\NDL{a} \subsetneq \NDL{b}.\] 
+\end{theorem}
+\begin{proof}
+
+\end{proof}
+
+Note that this theorem relies heavily on the assumption that $\SAT\in\NDL{c}$,
+hence this theorem is \textit{conditional}.
+It is an open problem to give an unconditional depth hierarchy theorem that 
+separates $\NDL{a}$ from $\NDL{b}$ for any $a < b$.
+
+%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Structure of optimal proofs}
-How to get the contradiction given the slowdown rule for $c < 4\cos(\pi /7)$.
+In this section, we describe
+the process that helps us build a proof, formed of successive applications
+of the speedup and slowdown rules, 
+that reaches a contradiction from the assumption
+$\SAT\in\NDL{c}$ for some $c < 4\cos(\pi /7)$.
+
+\begin{observation}
+	Consider a quantifier trading proof starting from $\ldots (Q_k~n^{a_k})\NDL{d}$
+	and that never goes below $k$ quantifiers.
+	Then, for the resulting class $\ldots (Q_k~n^{a_k})\NDL{d'}$, we have $d' \ge c\cdot a_k$.
+\end{observation}
+This is due to the presence of $a$ in the $\max$ in the Slowdown 
+rule (Corollary~\ref{cor:extended_slowdown}).
+In what follows, we aim to reach $d' = c\cdot a_k$ 
+and want to apply this process inductively,
+therefore we will start from depth $d = \frac{c^2}{2}a_{k+1}\log n$
+(which result from a slowdown on depth $ca_{k+1}\log n$).
+Not all values can be reduced to $c\cdot a_k$: 
+a study of necessary and sufficient conditions
+will give us a lower and an upper bound on $a_{k+1}$ in terms of $c$ and $a_k$,
+from which we will derive an explicit construction of an optimal choice of values for $(a_k)_k$.
+
+%%%%%%%
+\paragraph{Reducing the depth to $c a_k \log n$}
+The critical part is to manage to reduce the depth of the verifier
+from $\frac{c^2}{2}\cdot a_{k+1} \log n$ to $ca_k \log n$,
+while preserving the last quantifier exponent to $a_k$.
+One way to do this is by applying a sequence of speedup-slowdown alternation.
+One application of speedup then slowdown proves the following:
+\begin{flalign*}
+	\ldots (Q_k~n^{a_k})\NDL{d} 
+		&\subseteq \ldots \left(Q_k~n^{\max(a_k, x)}\right) (Q_{k+1}~n) 
+			\NDL{(d-2x)} \text{ Speedup w/param.} x\\ 
+		&\subseteq \ldots \left(Q_k~n^{\max(a_k, x)}\right) 
+			\NDL{c\cdot\max(a_k, x, (d-2x)/2, 1)} \text{ Slowdown} 
+\end{flalign*}
+Since we do not want to increase the exponent of the $Q_k$ quantifier,
+we get the following constraint:
+\begin{equation}
+	x \le a_k \label{eq:cstr1}
+\end{equation}
+Moreover, assuming that $a_k$ is at least $1$ allows us to reformulate
+the last class as 
+\[\ldots \left(Q_k~n^{a_k}\right) \NDL{c\cdot\max(a_k, (d-2x)/2)}.\]
+
+From here, we can get the following:
+\begin{lemma}[``Squiggle rule'']\label{lemma:squiggle}
+	Let $a,d$ be positive real numbers such that $ac < d < \frac{2ca}{c-2}$.
+	Then, the proof $(1_{a}0)^t, t \le \left\lceil\frac{d}{(4-c)a}\right\rceil + 1$
+	proves that
+	\[\ldots (Q~n^a)\NDL{d} \subseteq \ldots (Q~n^a)\NDL{ca}.\]
+\end{lemma}
+\begin{proof}
+	If $d/2 - a \le a \Leftrightarrow d \le 4a$,
+	then we get the desired result.
+	Setting $x = d/2 - a_k$ would immediately yield the desired depth.
+
+	We now assume that $d > 4a$, and show that we can get to $d \le 4a$ in a finite number of steps.
+	When $d \ge 4a$, applying the speedup rule with parameter $x = a$
+	followed by the slowdown rule transforms
+	$d$ is into $c\cdot(d/2 - a)$.
+
+	This is an improvement whenever $d$ is such that:
+	\begin{equation}
+		d < \frac{2ca_k}{c-2} \label{eq:cstr3}
+	\end{equation}
+	Moreover, as long as $c < 4$, we have $\frac{2c}{c-2} > 4$, hence 
+	there is a gap between $4a$ and $\frac{2ca}{c-2}$ where we can get a non-trivial improvement.
+
+	The improvement is $d(1 - c/2) + ca > (4-c)a$, since we are in the case $d > 4a$.
+	Therefore, as the improvement is lower bounded by a constant,
+	iterating this process reaches a value below $4a$ 
+	starting from $d$ in $\left\lceil\frac{d}{(4-c)a}\right\rceil$ steps,
+	after which one more step with param $x = d/2 - a$
+	yields the desired $ca \log n$ depth.
+\end{proof}
+
+%%%%%%%
+\paragraph{Combining all the pieces}
+As shown above, we need that $d < \frac{2ca_k}{c-2}$ 
+to be able to reach $ca_k$ starting from $d$.
+Since we want to apply this argument inductively, we will have 
+$d = \frac{c^2}{2}a_{k+1}$.
+Combining the above yields the following:
+\begin{equation}
+	a_{k+1} < \frac{4a_k}{c(c-2)} \label{eq:cstr_ak}
+\end{equation}
+Notice that $\frac{4}{c(c-2)} < 1$ whenever $c > 2\phi = 1 + \sqrt{5}$, 
+where $\phi$ denotes the golden ratio.
+
+Eq.~\ref{eq:cstr_ak} leads us to the following definition.
+\begin{definition}
+	Let $c > 2\phi$.
+	A sequence of real number $(a_k)_{k=0,\ldots,t}$ is \textit{well-behaved for $c$}\todo{find a name for that}
+	if the following holds:
+	\begin{enumerate}
+		\item for every $k=0,\ldots,t, a_k \ge 1$,
+		\item $a_t = 1$,
+		\item for every $k=0,\ldots,t, c a_{k+1}/2  \ge a_k$, and
+		\item\label{constr:decr} for every $k=0,\ldots,t-1, a_{k+1} < \frac{4a_k}{c(c-2)}$.
+	\end{enumerate}
+\end{definition}
+Notice that constraint~\ref{constr:decr} above implies that $a_{k+1} < a_k$,
+as $c > 2\phi$ implies that $\frac{4}{c(c-2)} < 1$.
+
+\begin{lemma}
+	Let $c > 2\phi$, let $(a_k)_{k=0,\ldots,t}$ be a sequence well-behaved for $c$,
+	and let $d_0 = 2\left(1+\sum_{k=0}^t a_k\right)$.
+	Then there exists a alternation trading proof that shows
+	\[\NDL{d_0} \subseteq \NDL{\frac{c^2}{2}a_0}.\]
+\end{lemma}
+\begin{proof}
+	The proof starts by applying $t$ speedups, with respective parameters $a_1,\ldots, a_t$.
+	We obtain the following:
+	\begin{flalign*}
+		\NDL{d_0}
+			&\subseteq (\exists n^{a_0}) (\forall n) \NDL{d_0 - 2_a0}\\
+			&\subseteq (\exists n^{a_0}) (\forall n^{a_1}) (\exists n) \NDL{d_0 - 2a_0 - 2a_1}\\
+			&\subseteq \vdots\\
+			&\subseteq (\exists n^{a_0}) \ldots (Q_t n^{a_t}) (Q_{t+1} n) \NDL{d_0 - 2\sum_{k=0}^t a_k}\\
+			&= (\exists n^{a_0}) \ldots (Q_t n^{a_t}) (Q_{t+1} n) \NDL{2}
+	\end{flalign*}
+	Then, applying Slowdown rule yields the class
+	\[(\exists n^{a_0}) \ldots (Q_t n^{a_t}) \NDL{c\cdot a_t}.\]
+	We will now repeat $t$ successive applications of the Slowdown rule and Squiggle rule.
+	Informally, it will give us the following inclusions:
+	\begin{flalign*}
+			(\exists n^{a_0}) \ldots (Q_t n^{a_t}) \NDL{c\cdot a_t}
+			&\subseteq (\exists n^{a_0}) \ldots (Q_{t-1} n^{a_{t-1}}) \NDL{\frac{c^2}{2}\cdot a_t} \text{ (Slowdown)} \\
+			&\subseteq (\exists n^{a_0}) \ldots (Q_{t-1} n^{a_{t-1}}) \NDL{c\cdot a_{t-1}} \text{ (Squiggle)} \\
+			&\vdots \\
+			&\subseteq (\exists n^{a_0}) \NDL{c\cdot a_0} \\
+			&\subseteq \NDL{\frac{c^2}{2}a_0} \text{ (Slowdown)}
+	\end{flalign*}
+
+	This can be shown formally by induction, mainly using the fact that $(a_k)_k$
+	is well-behaved for $c$:
+	during the $i$-th step ($i = 0,\ldots, t-1$), we have the following properties:
+	\begin{enumerate}
+		\item\label{inv1} before applying the Slowdown rule,
+		the depth of the verifier is $c\cdot a_{t-i}\log n$,
+		\item\label{inv2} after applying the Slowdown rule,
+		the depth is $\frac{c^2}{2}a_{t-i}\log n$,
+		\item\label{inv3} after applying the Squiggle rule,
+		the depth is $c\cdot a_{t-i-1}\log n$.
+	\end{enumerate}
+	A step $i = t$, we only apply the Slowdown rule
+	(we cannot apply the Squiggle rule: there is no leading quantifier).
+
+	Property~\ref{inv1} is true for $i = 1$, and then holds by induction when Property~\ref{inv3}
+	holds for $i-1$.
+	Property~\ref{inv2} hold when Property~\ref{inv1} holds, 
+	by definition of the Slowdown rule (Corollary~\ref{cor:extended_slowdown}),
+	and using the fact that $a_i \ge 1$ and $ca_i > a_{i-1}$.
+	Finally, Property~\ref{inv3} hold when Property~\ref{inv2} holds:
+	we can apply the Squiggle rule (Lemma~\ref{lemma:squiggle}) since $a$ is well-behaved for $c$,
+	which ensures that $i=0,\ldots,t-1, a_{i+1} < \frac{4a_i}{c(c-2)}$.
+\end{proof}
+
+The last step is to, given a $c < 4\cos(\pi/7)$, construct a sequence well-behaved for $c$
+such that ${d_0} > \frac{c^2}{2}a_0$, which yields the desired contradiction.
+The following lemma constructs such a sequence explicitly.
+\begin{lemma}\label{lemma:construction}
+	Let $c \in ]2\phi, 4\cos(\pi/7)[$, and let $\tau = \frac{c(c-2)}{4}$.
+	Then there exists an integer $t$ such that the sequence
+	$(r\cdot\tau^{t-i} - \eps)_{k=0,\ldots, t}$ is well-behaved for $c$, where $r = 1 + \eps$,
+	and $d_0  = 2\left(1+\sum_{k=0}^t a_k\right)$ is greater than $\frac{c^2}{2}a_0$.
+\end{lemma}
+\begin{proof}
+	We first check that $(a_k)_k$ is well behaved for $c$.
+	Notice that since $r = 1 + \eps$, we have $a_t = 1$,
+	and since $\tau > 1$, we have $a_k \ge 1$ for every $k = 1,\ldots, t$.
+	Now, for every $k < t$, we have $a_k + \eps = \tau(a_{k+1} + \eps)$,
+	which, combined with $\tau > 1$ implies that
+	\[a_{k+1} = a_k/\tau +\eps(\tau^{-1} - 1) < \frac{a_k}{\tau} = \frac{4a_k}{c(c-2)}.\]
+	The constraint $ca_{k+1} \ge a_k$ can we rewritten as
+	\[a_{k+1} (\frac{c}{2} - \tau) \ge \eps (\tau -1).\]
+	Since $\eps = 1/(t+1)$, the RHS of this inequality goes to $0$ as $t$ goes to infinity,
+	and $(\frac{c}{2} - \tau) > 0$ when $c$ is greater than $3$ (which is the case here since $c > 2\phi$).
+	Therefore, there exists a $t$ such that the inequality holds for every $k$, 
+	and the sequence $(a_k)_{k=1,\ldots,t}$ is well-behaved for $c$.
+
+	We now show that there exists a $t$ such that $d_0 > \frac{c^2}{2}a_0$.
+	We have:
+	\begin{flalign*}
+	d_0 
+		&= 2\left(1 - (t+1)\eps + \sum_{k=0}^t r\tau^k\right)\\
+		&= 2\left(1 - 1 + r\sum_{k=0}^t \tau^k\right)\\
+		&= 2r\frac{\tau^{t+1}-1}{\tau-1}
+	\end{flalign*}
+	Therefore, we have
+	\begin{flalign*}
+		d_0 > \frac{c^2}{2}a_0 
+		&\Leftrightarrow 2r\frac{\tau^{t+1} -1}{\tau-1} > \frac{c^2a_0}{2} \\
+		&\Leftrightarrow r\frac{\tau\tau^t  - 1}{\tau - 1} > \frac{c^2(r\tau^t-\eps)}{4} \\
+		&\Leftrightarrow r\tau^t\left[\frac{c^2}{4} - \frac{\tau}{\tau - 1}\right] < \frac{-1}{\tau - 1} + \eps\frac{c^2}{4}\\
+		&\Leftrightarrow \frac{c^2}{4}\tau - \frac{c^2}{4} - \tau < (r\tau^t)^{-1}\left[- 1 + \frac{c^2(\tau - 1)}{4(t+1)}\right]\\
+		&\Leftrightarrow \frac{c^2}{4}\tau - \frac{c^2}{4} - \tau < \frac{-1}{(1+\eps)\tau^t} + \frac{c^2(\tau - 1)}{4(t+1)(1+\eps)\tau^t}
+	\end{flalign*}
+	We can rewrite the LHS of the last inequality as follows:
+	\begin{flalign*}
+		\frac{c^2}{4}\tau - \frac{c^2}{4} - \tau 
+			&= \frac{c^3(c-2)}{16} - \frac{c^2}{4} - \frac{c(c-2)}{4}\\	
+			&= \frac{1}{4}\left(\frac{c^4}{4} - \frac{c^3}{2} - 2c^2 + 2c\right)
+	\end{flalign*}
+	Moreover, the RHS of that same last inequality is negative
+	and goes to $0$ as $t$ goes to infinity, 
+	while the LHS does not depend on $t$.
+	Therefore, for every $\delta < 0$, there exists a $t$ such that the last inequality is satisfied
+	whenever we have:
+	\[c^3 - 2c^2 - 8c + 8 < \delta.\]
+	The polynomial $P = X^3 - 2X^2 - 8X + 8$ is negative whenever $X$ is between
+	its 2nd largest $r_2 \simeq 0.89$ and largest root $r_1 = 4\cos(\pi/7)$.
+
+	Now, fix some $c \in ]r_2, r_1[$, and let $\delta = P(c) / 2 > P(c)$.
+	From the above, there exists a $t$ such that $c^3 - 2c^2 - 8c + 8 < \delta$
+	implies $d_0 > \frac{c^2}{2}a_0$, and therefore
+	the proof for $(a_k)_k$ leads to a contradiction
+	assuming $\SAT\in\NDL{c}$.
+\end{proof}
+
+Lemma~\ref{lemma:construction} proves Theorem~\ref{thm:main} for $c \in]2\phi, 4\cos(\pi/7)[$;
+to extend this result to $c \le 2\phi$, it suffices to notice that in that case,
+we have $\NDL{c} \subseteq \NDL{(2\phi + \eps)}$, hence it reduces to the former case.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\section{Improving Brent/Spira}
-\todo[inline]{Prove this}
-Useful starting point : Size-Depth Tradeoffs for Boolean Formulae (Bonet, Buss).
+\section{Can we turn depth lower bounds into (good) size lower bounds?}
+Can we prove an improved Brent/Spira-type theorem that turns our lower bounds into super-cubic size lower bounds?
+Useful starting point: Size-Depth Tradeoffs for Boolean Formulae (Bonet, Buss).
 
-Remark: we must focus on the case when the input formula is a \NAND{} formula.
-Otherwise, we prove a much stronger result.
-Maybe prove the stronger result with a $2+\eps$ factor? Not very useful though.
-If we can go below 2, we improve something.
+%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%
+\section{Conclusion and future work}
+The lower bounds given in this work are against a very restricted
+type of formulas (NAND formulas, or equivalently alternating OR/AND formulas).
+An open problem is to lift these lower bounds to other computational bases,
+for example by allowing XOR and equality gates,
+using ``even'' or ``odd-parity'' quantifiers, 
+and assuming that \ParSAT{} has low-depth formulas.
+In that case, we can get a lower bound for \SAT{} \textit{or} \ParSAT{}.
+We can go further, and get lower bounds for \SharpSAT{},
+since \ParSAT{} is the least significant bit of \SharpSAT{}.
+The first idea to solve \SAT{} (that is, to remove existential quantifiers) 
+with \SharpSAT{} would be to take the OR of all $n+1$
+bits of \SharpSAT{}, but this add depth $\log n$ to the formula.
+However, \textsf{Tautology} can be solved
+with the most significant bit of \SharpSAT{}:
+therefore we may be able to remove existential quantifiers 
+by looking at a single bit of \SharpSAT{}
+for a formula that corresponds to a negation of the verifier.
+
+We may also be able to obtain results against less-restricted formulas,
+for example by allowing two types of gates within a layer,
+while keeping an alternating structure.
+For example: odd-numbered layers are ANDs and ORs mixed,
+even-numbered layers are XORs and EQs mixed.
+
+We can also relax the structure of the formulas: 
+instead of alternating layers, we can only ask
+that at least one layer in every group of $k$ layers
+is an OR layer, and at least one other is an AND layer.
+We get weaker speedups and slowdowns
+(namely, the $1/2$ in the exponent becomes $\alpha_k = (k-1)/k$),
+but against more expressive formulas.
 
@@
 \section{Program to find optimal parameters}\label{app:ilp}
+
+\subsection{Overview of the approach}
 Improves over Williams (cite): no need to enumerate proof annotations.
@@
 Therein, Williams enumerates all possible proof skeletons of length $k$,
-and tries to find to find the best set of parameters for each of them, using a linear program.
+and tries to find the best set of parameters for each of them, using a linear program.
 In this work, we propose a different approach, that proves to be more efficient in practice:
@@
 More precisely, we give an ILP $P_{c,k}$ which is feasible if and only if
-there exists a proof annotation $p$ of length $k$\todo{it's not exactly that, it's $k$ up steps} along with parameters $x, r_0$
-such that $p$ proves a contradiction under the assumption that $\SAT\in\NDL{c}$.
+there exists a proof annotation $p$ of length $k$,\footnote{Our program takes as input a value $k'$ and searches over proofs with $k'$ application of the speedup rule, which implies $k'+1$ slowdowns since every quantifier must be removed to reach a contradiction.
+Therefore, we have $k = 2k'+1$.} along with parameters $x, r_0$
+such that $p$ proves a contradiction under the assumption that $\SAT\in\NDL{c}$
 Moreover, its solution encodes $p,q,r_0$.
 By using binary search over $c$, we can quickly get an approximation 
-of the largest $c$ such that a proof of length $k$ can reach a contradiction under $H_c$\todo{define "hypothesis c"}, together with a proof for $c-\eps$.
-Repeating the above steps for increasing values of $k$ shows that the sequence $c_k^*$\todo{define ?}
-Table of $c_k^*$ for first few values of $k$?
-Seems to show convergence for some value $c^\infty > 3.6$.
+of $c_k^*$, the largest $c$ such that a proof of length $k$ can reach a contradiction assuming that $\SAT\in\NDL{c}$.
+Repeating the above steps for increasing values of $k$ shows that the sequence $(c_k^*)_{k}$
+seems to converge to some value $c_\infty^* > 3.5$ (see Table~\ref{table:ilp_res}).
 
-\todohere describe model + link to code?
+\begin{table}[htbp]
+	\begin{center}
+	\begin{tabular}{ |c|l|l| } 
+	\hline
+	$k$ & $c_k^*$ & Skeleton of $p_k^*$ \\ 
+	\hline
+	3 & 2.82812 & \texttt{100} \\
+	7 & 3.20020 & \texttt{1100100} \\
+	11 & 3.32617 & \texttt{11100100100} \\
+	15 & 3.39844 & \texttt{111001010010100} \\
+	19 & 3.44629 & \texttt{1111001001010010100} \\
+	23 & 3.47559 & \texttt{11110010100101001010100} \\
+	27 & 3.49805 & \texttt{111110010010100101001010100} \\
+	\hline
+	\end{tabular}
+	\end{center}
+	\caption{Approximate value of $c_k^*$ and skeleton of $p_k^*$ for some values of $k$.}\label{table:ilp_res}
+\end{table}
+
+\subsection{Details of the model}
+\todoh{describe model}
+\todoh{link to code?}
 

### 2022-06-26 article/main.tex
Added defs to intro, done

@@
 The depth of a formula is the depth of the underlying tree.
+Given an input $x \in\bit^n$, the value of $\phi$ at $x$ can be defined 
+as the value of it root on input $x$, 
+where the value of a gate is defined inductively as follows:
+\[val(g) =\begin{cases}
+	c & \text{ if $g$ is labeled with the constant} c\in\bit\\
+	x_i & \text{ if $g$ is labeled with $x_i$}\\
+	\neg x_i & \text{ if $g$ is labeled with $\neg x_i$}\\
+	f(g_1, g_2) & \text{ if $g$ is labeled  $f$, and $g_1,g_2$ are children of $g$}
+\end{cases}\]
 
@@
 
-\todoh{To introduce: family of formulas computing a language}
-% In this work, we focus on formulas with depth $d(n) = (c + o(1))\log n$, for some $c > 1$.
-However, there exists formulas of \textit{constant} depth that compute undecidable languages.
+A family of formulas is a sequence $(\phi_n)_{n\in\NN}$ of formulas such that
+for every i, $\phi_i$ is a formula over $i$ inputs.
+A language $L\subseteq \bit^*$
+is computed by\footnote{In what follows, we may use ``$L$ has formulas ...'' for ``$L$ is computed by a family of formulas ...''} 
+the family of formulas $(\phi_n)_{n\in\NN}$
+if for every $n$ and every $x\in\bit^n$, $\phi_n(x) = 1 \Leftrightarrow x\in L$.
+
+However, there exists families formulas of \textit{constant} depth that compute undecidable languages.
 In this work, we restrict our attention to uniform formulas, 
@@
 \subsection{Details of the model}
+Given an integer $k$ and a real $c$, we define the following interger linear program $P_{c,k}$ as follows:
+\begin{equation}
+	todo
+\end{equation}
 \todoh{describe model}

### 2022-06-27 article/main.tex
Strange mistake in LP

@@
 \usepackage[margin=3cm]{geometry}
+\usepackage{hyperref}
 \usepackage{amsfonts}
@@
 \usepackage{enumitem}
-\usepackage{algorithm}
-\usepackage[noend]{algpseudocode}
+% \usepackage{algorithm}
+% \usepackage[noend]{algpseudocode}
 
@@
 		&\subseteq \ldots (\forall n^{a}) (\exists n^b) \NTIME[n^{d/2}]\\	
-		&\subseteq \ldots (\forall n^{a}) \NTIME[n^{\max(d/2, d)}]	
+		&\subseteq \ldots (\forall n^{a}) \NTIME[n^{\max(d/2, b)}]	
 	\end{flalign*}
@@
 \subsection{Details of the model}
-Given an integer $k$ and a real $c$, we define the following interger linear program $P_{c,k}$ as follows:
-\begin{equation}
-	todo
-\end{equation}
-\todoh{describe model}
-\todoh{link to code?}
+Given an integer $k \ge 3$ and a real $c$, we define the interger linear program $P_{c,k}$ as follows:
+\[
+\begin{array}{rrcll}
+	(P_{k,c}) &\max& 1&&\\
+	\text{Initial speedup constraints:}&&&&\\
+	&d_1&=&d_0 - 2x_1&\\
+	&a_{1, -1}&=&1&\\
+	&a_{1, -2}&\ge&x_1&\\
+	&su_1&=&1&\\
+	\text{Speedup constraints:}&&&&\\
+	\forall i \ge 1, su_i = 1\Rightarrow &d_i&=&d_{i-1} - 2x_i&\\
+	\forall i \ge 1, su_i = 1\Rightarrow &a_{i, j}&\ge&a_{i-1, j+1}& j = 1,\ldots,k-1\\
+	\forall i \ge 1, su_i = 1\Rightarrow &a_{i, -2}&\ge&x_i&\\
+	\forall i \ge 1, su_i = 1\Rightarrow &a_{i, -1}&=&1&\\
+	\text{Slowdown constraints:}&&&&\\
+	\forall i \ge 1, su_i = 0\Rightarrow &d_i&\ge&c\cdot d_{i-1}/2&\\
+	\forall i \ge 1, su_i = 0\Rightarrow &d_i&\ge&c\cdot a_{i-1, -1}&\\
+	\forall i \ge 1, su_i = 0\Rightarrow &d_i&\ge&c\cdot a_{i-1, -2}&\\
+	\forall i \ge 1, su_i = 0\Rightarrow &a_{i, j}&=&a_{i-1, j-1}& j = 2,\ldots,k\\
+	\text{Well-formed proof constraints:}&&&&\\
+	\forall i \ge 1, & 2\sum_{t=1}^i su_{t} &\ge& i&\\ 
+					 & 2\sum_{i=1}^k su_{i} &=& k - 1&\\ 
+	\text{Require contradiction:}&&&&\\
+	& d_k &<&d_0&\\
+	\text{Variables:}&&&&\\
+	&d_i &\in& [0,+\infty), &i =0,\ldots, k\\
+	&a_{i,j} &\in& [1,+\infty), &i,j =1,\ldots, k\\
+	&x_i &\in& [0,+\infty), &i =1,\ldots, k\\
+	&su_{i} &\in& \{0,1\}, &i =1,\ldots, k
+\end{array}
+\]
+The variables have the following meaning:
+\begin{enumerate}
+	\item $d_i$: verifier is $\NDL{d_i}$ on line $i$,
+	\item $a_{i,k+1-j}$: exponent of the $j$-th-to-last quantifier on line $i$. 
+	We use
+	$a_{i,-j}$ as a shorthand notation for $a_{i,k+1-j}$.
+	\item $x_i$: if line $i$ is the result of a speedup, it is a speedup with parameter $x_i$,
+	\item $su_{i}$: 1 if and only if line $i$ results in the application of a speedup, 0 otherwise.
+\end{enumerate}
+The constains of the type ``$su_i = b$ implies $C$'' can be encoded using the so-called ``Big M method''\footnote{See \url{https://en.wikipedia.org/wiki/Big\_M\_method}.}. 
+The constaint $d_k < d_0$ can be approximately encoded as $d_k \le d_0 - \delta$ for some small $\delta > 0$.
+
+The proof leads to a contradiction if and only if $d_k < d_0$, 
+that is, if and only if the ILP is feasible.
+Using binary search on $c$, we can quickly find an upper and lower bounds on $c_k^*$.
 

### 2022-06-27 with_mip/main.py

@@ -43,13 +43,12 @@ def run_proof(annotation, params, p0, c):
     for i in range(1, k):
         if annotation[i] == '1':
             x = params[i+1]
-            q.append(q[-1])
+            q.append(1)
             q[-2] = max(q[-2], x/2)
             r = r - x
             print_anno(q, r, x)
         elif annotation[i] == '0':
-            x = params[i+1]
-            r = c*max(r-x, x, q[-1], q[-2]) if len(q) > 1 else c*max(r-x, x, q[-1])
+            r = c*max(r/2, q[-1], q[-2]) if len(q) > 1 else c*max(r/2, q[-1], 1)
             q.pop()
             print_anno(q, r, None)
         elif annotation[i] == '2':
@@ -75,7 +74,6 @@ def time_generated_search(size, f, print_proof=False, **kwargs):
     print(f'Best c found: {best_c:.5f} in {(t2-t):.5f}s for annotation {best_proof}')
     if print_proof:
         run_proof(best_proof, best_x, best_r, best_c)
-    
 
 def time_search(size, f, print_proof=False, **kwargs):
     t = time()
@@ -85,30 +83,8 @@ def time_search(size, f, print_proof=False, **kwargs):
     if print_proof:
         run_proof(best_proof, best_x, best_r, best_c)
 
-def good_annotation(k):
-    if k == 0:
-        return '0'
-    return '1' + good_annotation(k-1) + '10'*k + '0'
-
-def try_good_proof():
-    for s in [3,5,10,15,30,50]:
-        t = time()
-        c,p,x,r = best_param_sparse(good_annotation(s))
-        t2 = time()
-        print(f'Best c found: {c:.5f} in {(t2-t):.5f}s for annotation {p}')
-
-
-def try_proof(p):
-    t = time()
-    c,p,x,r = best_param_sparse(p, verbose=1)
-    t2 = time()
-    print(f'Best c found: {c:.5f} in {(t2-t):.5f}s for annotation {p}')
-
-
 if __name__ == '__main__':
-    # try_good_proof()
-    # try_proof('11010100100')
-    # exit(0)
+
     parser = argparse.ArgumentParser()
     parser.add_argument("depth", help="Number of 'up' steps in target proofs", type=int)
     parser.add_argument('-v', '--verbose', help="Activate debug of search functions", action='count', default=0)

### 2022-06-27 with_mip/pysearch.py

@@ -21,21 +21,20 @@ def best_proof_sparse(k, c_lb=1.0, c_ub=5.0, tol=0.001, n_threads=10, verbose=0)
         m.Params.Threads = n_threads
         nl = k+1 # number of lines
         nc = k//2+1+5 # number of columns
-        a = m.addVars(nl, nc, name='a')
+        a = m.addVars(nl, nc, lb=1, name='a')
         x = m.addVars(nl, name='x') # params for speedups
         r = m.addVars(nl, name='r') # runtime
         su = m.addVars(nl, vtype=GRB.BINARY, name='su') # usual speedup
         sd  = m.addVars(nl, vtype=GRB.BINARY, name='sd')  # slowdown
 
         """
-        We proceed differently: we put the last quantifier of line i in a[i, -1], 
+        We proceed differently: 
+        we put the last quantifier of line i in a[i, -1], 
         2nd to last in a[i, -2], etc.
         """
 
         # Apply the first speedup
         m.addConstr(a[1, nc-2] >=  x[1]/2) 
-        m.addConstr(a[1, nc-2] >=  1) 
-        m.addConstr(a[1, nc-1] >=  1)
         m.addConstr(r[1] ==  r[0] - x[1])
         m.addConstr(su[1] == 1)
         
@@ -46,7 +45,6 @@ def best_proof_sparse(k, c_lb=1.0, c_ub=5.0, tol=0.001, n_threads=10, verbose=0)
         m.addConstrs((su[i] == 1) >> (a[i, j] >= a[i-1, j+1]) for j in range(nc-1) for i in range(2, nl))
         #   Set speedup arg "max(x/2, a)"
         m.addConstrs((su[i] == 1) >> (a[i, nc-2] >= x[i]/2)       for i in range(2, nl)) 
-        m.addConstrs((su[i] == 1) >> (a[i, nc-1] >= a[i-1, nc-1]) for i in range(2, nl)) 
         #   Set runtime
         m.addConstrs((su[i] == 1) >> (r[i] == r[i-1] - x[i]) for j in range(1, nc) for i in range(2, nl)) 
  
@@ -56,16 +54,14 @@ def best_proof_sparse(k, c_lb=1.0, c_ub=5.0, tol=0.001, n_threads=10, verbose=0)
         #       in order for the proof to be "well-typed".
         #   Copy l-1 quantifiers
[…]

### 2022-06-27 with_mip/pyparams.py
Strange mistake in LP

@@ -73,7 +73,7 @@ def best_param_sparse(proof, c_lb=1.0, c_ub=5.0, tol=0.001, verbose=0):
             elif r_type == '1': # speedup
                 m.addConstr(a[i] >= stack[-1])
                 m.addConstr(a[i] >= x[i]/2)
-                m.addConstr(b[i] >= stack[-1])
+                m.addConstr(b[i] >= 1)
                 m.addConstr(r[i] == r[i-1] - x[i])
                 stack.pop()
                 stack.append(a[i])

### 2022-06-27 with_mip/main.py
Strange

@@ -37,7 +37,7 @@ def run_proof(annotation, params, p0, c):
     print()
     print(f'Nand-depth[{p0} log n]')
     x = params[1]
-    q = [x/2, 1]
+    q = [max(x/2, 1), 1]
     r = p0 - x
     print_anno(q, r, x)
     for i in range(1, k):
@@ -48,13 +48,9 @@ def run_proof(annotation, params, p0, c):
             r = r - x
             print_anno(q, r, x)
         elif annotation[i] == '0':
-            r = c*max(r/2, q[-1], q[-2]) if len(q) > 1 else c*max(r/2, q[-1], 1)
+            r = c*max(r/2, q[-1], q[-2], 1) if len(q) > 1 else c*max(r/2, q[-1], 1)
             q.pop()
             print_anno(q, r, None)
-        elif annotation[i] == '2':
-            r = c*max(r, q[-1], q[-2]) if len(q) > 1 else c*max(r, q[-1])
-            q.pop()
-            print_anno(q, r)
         else:
             print('Incorrect annotation')
             exit(0)
@@ -65,11 +61,14 @@ from time import time
 
 def time_generated_search(size, f, print_proof=False, **kwargs):
     t = time()
-    best_c, best_proof, best_x, best_r = 0, '', [], 0
+    best_c, best_proof, best_x, best_r = 0, [], [], 0
     for p in enumerate_annotations(size):
         c, proof, x, r = f(p, **kwargs)
         if c > best_c:
-            best_c, best_proof, best_x, best_r = c, proof, x, r
+            best_c, best_proof, best_x, best_r = c, [proof], x, r
+        elif c == best_c:
+            best_proof.append(proof)
+            print(c, best_proof)
     t2 = time()
     print(f'Best c found: {best_c:.5f} in {(t2-t):.5f}s for annotation {best_proof}')
     if print_proof:

### 2022-06-27 with_mip/pysearch.py

@@ -100,7 +100,7 @@ def best_proof_sparse(k, c_lb=1.0, c_ub=5.0, tol=0.001, n_threads=10, verbose=0)
                     best_r = r[0].x
                     if verbose > 0:
                         print(f'Value {mid} feasible')
-                        pretty_print(a, nl, nc)
+                        # debug(a, nl, nc)
             else:
                 c_ub = mid
                 if verbose > 0:
@@ -113,6 +113,6 @@ def best_proof_sparse(k, c_lb=1.0, c_ub=5.0, tol=0.001, n_threads=10, verbose=0)
     except AttributeError:
         print('Encountered an attribute error')
 
-def pretty_print(a, nl, nc):
+def debug(a, nl, nc):
     for i in range(nl):
         print(' '.join(f'(Q n^{a[i,j].x})' for j in range(nc)))

### 2022-06-27 article/main.tex
Report is done

@@
 \usepackage{enumitem}
-% \usepackage{algorithm}
-% \usepackage[noend]{algpseudocode}
 
@@
 \newtheorem{theorem}{Theorem}[section] % Number within sec ?
-% \newtheorem{theorem}{Theorem}
 \newtheorem{prop}[theorem]{Proposition}
@@
 \newtheorem{remark}[theorem]{Remark}
-% \newtheorem*{remark*}{Remark}
-% \theoremstyle{definition}
-% \newtheorem{problem}{Problem}
-% \newtheorem{openproblem}{Open Problem}
 
@@
 
-\newcommand{\todohere}{\todo[inline]{TODO}}%
-\newcommand{\todoh}[1]{\todo[inline]{TODO: #1 }}%
-
 %%%%%%%%%%%%%%%%%%%%%%
@@
 
-\todoh{Motivation: understanding complexity. Explicit lower bounds: programs, circuits, formulas.
-Uniformity.}
+As a first step to understand the relations between different complexity classes,
+an important problem in complexity theory is to find
+lower bounds lower bounds for explicit problems in various model: programs, circuits, or formulas.
+
+A particular attention was dedicated to the \SAT{} problem, 
+since a super-polynomial lower bound for this problem would imply $\NP \neq \P$
+(for programs, $\NP \neq \textsf{P/poly}$ for circuits).
 
 A line of work initiated by Fortnow~\cite{fortnow2000time,fortnow2005time} in 1997 
-has proven model-independent time-space lower bounds for the SAT
-problem and related problems within the polynomial-time hierarchy.
+has proven model-independent time-space lower bounds
+(or equivalently, lower bounds against small-space algorithms) for \SAT{}
+and related problems within the polynomial-time hierarchy.
 This approach was then extended by Williams~\cite{williams2006inductive,williams2007time,williams2013alternation} 
@@
 the set of gates that are allowed in the formula.
-For the full binary basis $B_2$, Nechiporuk\cite{nechiporuk66boolean}
+For the full binary basis $B_2$, Nechiporuk~\cite{nechiporuk66boolean}
 proved in 1966 an $\Omegat{n^2}$ lower bound for an explicit function in \NP{}, 
@@
 For the De Morgan basis $\{\wedge, \vee, \neg\}$, Andreev~\cite{andreev1987method} 
-proved an $\Omegat{n^{2.5}}$ lower bound, 
+proved an~$\Omegat{n^{2.5}}$ lower bound, 
 which was later improved by H{\aa}stad~\cite{hastad1998shrinkage} to $\Omegat{n^{3}}$.
 This lower bound is optimal up to polylog factors,
-since their function has (uniform) formulas of size $\Ot{n^{3}}$.
+since this function has (uniform) formulas of size $\Ot{n^{3}}$.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\subsection{Results overview}
+\paragraph{Results overview}
 
-The \NAND{} basis is rather expressive: Andreev's function has \NAND{} formulas of depth $(3+o(1)) \log n$ (and therefore of size $\Ot{n^3}$), which matches the lower bound for the De Morgan Basis.
-\todoh{more about nand?}
+In this work, we focus in the \NAND{} basis, 
+that is formulas where every gate computes the function $\NAND: x,y\mapsto \neg(x\wedge y)$;
+there are no internal, only inputs can be negated.
+The \NAND{} basis is universal (i.e. can compute any boolean function), 
+and is rather expressive: Andreev's function has \NAND{} formulas of depth~$(3+o(1)) \log n$
+(and therefore of size $\Ot{n^3}$), which matches the lower bound for the De Morgan Basis.
 
-Namely, we prove the following:
+Our main result is the following:
 \begin{theorem}\label{thm:main}
@@
 \end{theorem}
+
+It is based on a novel algorithm that computes the 
+value of a \NAND{} formula in $\Sigma_2$ sublinear time.
+Our algorithm uses the equivalence between \NAND{} formulas and alternating OR/AND formulas.
 
@@
 whose internal nodes, called gates, are labeled by a function $f: \bit^2 \rightarrow \bit$,
-and leaves are labeled with a variable, $x_i$ its negation $\neg x_i$ for some $i, 1\le i \le n$,
+and leaves are labeled with a variable~$x_i$, its negation $\neg x_i$ for some $i, 1\le i \le n$,
 or a constant $c\in\bit$.
@@
 is computed by\footnote{In what follows, we may use ``$L$ has formulas ...'' for ``$L$ is computed by a family of formulas ...''} 
-the family of formulas $(\phi_n)_{n\in\NN}$
+the family of formulas~$(\phi_n)_{n\in\NN}$
 if for every $n$ and every $x\in\bit^n$, $\phi_n(x) = 1 \Leftrightarrow x\in L$.
@@
 \end{lemma}
-
+\begin{proof}
 	The intuitive idea behind this lemma is that the only
@@
 	therefore we only need to guess $O(n^{x/2})$ bits. 
+	Let $y$ be the corresponding bitstring.
 
@@
 	Note that while the second quantifier only requires $O(\frac{x}{2}\log n)$ nondeterministic bits,
-it still computes and outputs $n^{1+o(1)}$ bits, hence the $n^1$ exponent.
+	it still computes and outputs $n + x\log n$ bits 
+	(the output along with 1 bit for each of the first $x\log n$ levels), hence the $n^1$ exponent. 
+	Let $z$ be the corresponding bitstring.
 
@@
 
-\begin{proof}[Proof of Lemma~\ref{lemma:speedup}]
-	We prove the result for $\exists\forall$ computations, 
-	the result for $\forall\exists$ computations follows from the fact
-	that $\NDL{d}$ is closed under taking the complement.
+	Consider the machine $M'$ that,
+	on input $a'$ of length $n' = n + x\log n$ for some $n$,
+	let $a' = ab$, where $a$ has length $n$ and $b$ has length $x\log n$,
+	and $M'$ returns $M'(a', i, n') = M(a, bi, n)$.
+	By construction, $M'$ describes $\NDL{d-x}$-formulas.
+	Moreover, if the original formula was true on $x$, 
+	then there exists a choice of $y$ such that for all $z$,
+	the formula described by $M'$ is true on input $xb(y,z)$,
+	where $b(y,z)$ is the bitstring that describes a path 
+	encoded by $y$ restricted to the choices of $z$.
+	This is an $\exists\forall$ computation with $M'$ as its last stage that accepts $L$.
 
-	Let $L\in\NDL{d}$, let $M : (x,i,n) \mapsto g$ be 
-	a descriptor machine for $L$, and let $x$ be an input of length $n$.
-
-	We now construct a machine $M'$ that describes $\NDL{d-x}$-formulas,
-	such that an $\exists\forall$ computation with $M'$ as its last stage accepts $L$.
-	On an input $x$
-
-	We describe how to build bit strings $y$ of length $n^{x/2}$ and
-	\todoh{finish details of proof}
+	The result for $\forall\exists$ follows since $\NDL{d}$ is closed under complement.
 \end{proof}
@@
 \begin{corollary}\label{cor:speedup}
-	For every $d > 0$, we have
+	For every $d \ge 1$, we have
 	\[\NDL{d} \subseteq \NTIME[n^{d/2}] \cap \coNTIME[n^{d/2}].\]
@@
 \begin{theorem}
-	If there exists $c > 0$ such that $\SAT \subseteq \NDL{c}$, 
+	If there exists $c > 0$ such that $\NTIME[n] \subseteq \NDL{c}$, 
 	then for all $b > a > 0$, we have
@@
 \begin{proof}
+	We give a proof by contradiction. Assume that there exists $0 < a < b$
+	such that $\NDL{b} \subseteq \NDL{a}$. 
+	A standard amplification by padding argument shows that this implies
+	$\NDL{d} \subseteq \NDL{a}$ for \textit{any} $d$.
+	More precisely, we show that $\NDL{r\cdot a} \subseteq \NDL{a}$
+	implies  $\NDL{r^2\cdot a}\subseteq \NDL{r\cdot a} \subseteq \NDL{a}$ for any $r > 1$.
+	Taking $r = b/a > 1$ and applying this result $t = \log_2\log_r d/a$ times, we get.
+	$\NDL{r^{2^t}} = \NDL{d} \subseteq \NDL{a}$.
 
+	Now, assuming that there exists $c \ge 1/2$ such that $\NTIME[n] \subseteq \NDL{c}$,
+	we have:
+	\begin{flalign*}
+		\NTIME[2a] 
+			&\subseteq \NDL{2ca}\\
+			&\subseteq \NDL{a}\\
+			&\subseteq \NTIME[a]
+	\end{flalign*}
+	The first inclusion follows from a padding argument, 
+	the second one follows from the amplification argument above as $2c > 1$,
+	the last inclusion uses efficient simulation of $\ND$ by RAMs.
+	The proof for $c < 1/2$ is similar, except that the second inclusion follows
+	from $2c < 1$.
+	But this proves $\NTIME[2a] \subseteq \NTIME[a]$ for some $a > 0$,
+	which contradicts the Nondeterministic Time Hierarchy Theorem.
 \end{proof}
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\subsection{Structure of optimal proofs}
+\subsection{Structure of asymptotically optimal proofs}
 In this section, we describe
-the process that helps us build a proof, formed of successive applications
-of the speedup and slowdown rules, 
+the process that helps us build an quantifier-trading proof, 
+formed of successive applications
+of the speedup and slowdown rules along with their parameters, 
 that reaches a contradiction from the assumption
 $\SAT\in\NDL{c}$ for some $c < 4\cos(\pi /7)$.
+While these proofs might not be the shortest that yield a contradiction
+for a given $c$, they present a very simple inductive structure that makes them easier to analyze.
+
+We introduce here the \textit{proof annotation} notation of Williams~\cite{mudigonda2020time},
+which succinctly describes a quantifier-trading proof: 
+let $1_x$ denote an application of the speedup rule with parameter $x$, 
+and let $0$ denote an application of the slowdown rule.
+The proof annotation of a quantifier-trading proof is the string corresponding
+to the concatenation of the representation of each of its steps.
+For example, the proof annotation corresponding to 
+the short quantifier-trading proof presented in Section~\ref{sec:overview}
+is $1_200$.
 
@@
 	Let $c > 2\phi$.
-	A sequence of real number $(a_k)_{k=0,\ldots,t}$ is \textit{well-behaved for $c$}\todo{find a name for that}
+	A sequence of real number $(a_k)_{k=0,\ldots,t}$ is \textit{well-behaved for $c$}
 	if the following holds:
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\section{Can we turn depth lower bounds into (good) size lower bounds?}
-Can we prove an improved Brent/Spira-type theorem that turns our lower bounds into super-cubic size lower bounds?
-Useful starting point: Size-Depth Tradeoffs for Boolean Formulae (Bonet, Buss).
+\section{Conclusion and future work}
+As mentioned in the introduction, the lower bounds 
+proved in this work are \textit{depth} lower bounds,
+which are weaker than size lower bounds, since the latter directly imply the former.
+However, there exists theorems that give a converse, 
+but the depth is reduced by a constant factor
+that depend on the type of formula.
+This follows from work of Spira~\cite{spira1971time}, Brent~\cite{brent1974parallel},
+and more recently Bonet and Buss~\cite{bonet1994size}.
+For NAND formulas, the best known theorems have constants that are greater than $2$,
+therefore using them does not improve over the naive $2\cos \pi/7$
+that follow from TM lower bounds.
+However, for most families of formulas, the constants in the best theorems
+are not known to be tight\footnote{See Sergeev~\cite{sergeev2019relation} (in russian) for a recent exposition 
+best upper and lower bounds known on the constants of multiple classes of formulas.}, 
+therefore, it might be possible to improve them
+and use our result to prove nontrivial size lower bounds,
+and maybe super-cubic size lower bounds.
 
-%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion and future work}
 The lower bounds given in this work are against a very restricted
@@
 
+In this section, we describe how we extend and improve 
+the approach of Williams~\cite{williams2006inductive, williams2007time} 
+to programmatically find the proof annotations of a given $c$ that yield a contradiction
+for the largest possible $c$.
+
 \subsection{Overview of the approach}
-Improves over Williams (cite): no need to enumerate proof annotations.
+We call a \textit{proof skeleton} a proof annotation stripped of its speedup parameters.
+For example, the proof skeleton of the annotation $1_200$ is $100$.
 
-Therein, Williams enumerates all possible proof skeletons of length $k$,
+In his work, Williams enumerates all possible proof skeletons of length $k$,
 and tries to find the best set of parameters for each of them, using a linear program.
 In this work, we propose a different approach, that proves to be more efficient in practice:
-we use an Integer Linear program whose solution is
-an optimal proof of length $k$
-along with its parameters.
+we use an Integer Linear Program (ILP) whose solution is
+an optimal proof annotation of length $k$.
 
@@
 there exists a proof annotation $p$ of length $k$,\footnote{Our program takes as input a value $k'$ and searches over proofs with $k'$ application of the speedup rule, which implies $k'+1$ slowdowns since every quantifier must be removed to reach a contradiction.
-Therefore, we have $k = 2k'+1$.} along with parameters $x, r_0$
+Therefore, we have $k = 2k'+1$.} along with parameters $x, d_0$
 such that $p$ proves a contradiction under the assumption that $\SAT\in\NDL{c}$
-Moreover, its solution encodes $p,q,r_0$.
+starting from the class $\NDL{d_0}$.
+Moreover, its solution encodes $p,q,d_0$.
 By using binary search over $c$, we can quickly get an approximation 
-of $c_k^*$, the largest $c$ such that a proof of length $k$ can reach a contradiction assuming that $\SAT\in\NDL{c}$.
+of $c_k^*$, the largest $c$ such that a proof of length $k$ 
+can reach a contradiction assuming that $\SAT\in\NDL{c}$.
 Repeating the above steps for increasing values of $k$ shows that the sequence $(c_k^*)_{k}$
@@
 	\hline
-	3 & 2.82812 & \texttt{100} \\
-	7 & 3.20020 & \texttt{1100100} \\
-	11 & 3.32617 & \texttt{11100100100} \\
-	15 & 3.39844 & \texttt{111001010010100} \\
-	19 & 3.44629 & \texttt{1111001001010010100} \\
-	23 & 3.47559 & \texttt{11110010100101001010100} \\
-	27 & 3.49805 & \texttt{111110010010100101001010100} \\
+	3 & 2.82812 & 100 \\
+	7 & 3.20020 & 1100100 \\
+	11 & 3.32617 & 11100100100 \\
+	15 & 3.41797 & 111001001100100 \\
+	19 & 3.46094 & 1111001001001100100 \\
+	23 & 3.49902 & 11110010011001001100100 \\
+	27 & 3.52930 & 111100011000111001001100100 \\
 	\hline
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Details of the model}
-Given an integer $k \ge 3$ and a real $c$, we define the interger linear program $P_{c,k}$ as follows:
+Given an integer $k \ge 3$ and a real $c$, we define the integer linear program $P_{c,k}$ as follows:
 \[
@@
 	\text{Initial speedup constraints:}&&&&\\
-	&d_1&=&d_0 - 2x_1&\\
+	&d_1&=&d_0 - x_1&\\
 	&a_{1, -1}&=&1&\\
-	&a_{1, -2}&\ge&x_1&\\
+	&a_{1, -2}&\ge&x_1/2&\\
 	&su_1&=&1&\\
 	\text{Speedup constraints:}&&&&\\
-	\forall i \ge 1, su_i = 1\Rightarrow &d_i&=&d_{i-1} - 2x_i&\\
+	\forall i \ge 1, su_i = 1\Rightarrow &d_i&=&d_{i-1} - x_i&\\
 	\forall i \ge 1, su_i = 1\Rightarrow &a_{i, j}&\ge&a_{i-1, j+1}& j = 1,\ldots,k-1\\
-	\forall i \ge 1, su_i = 1\Rightarrow &a_{i, -2}&\ge&x_i&\\
+	\forall i \ge 1, su_i = 1\Rightarrow &a_{i, -2}&\ge&x_i/2&\\
 	\forall i \ge 1, su_i = 1\Rightarrow &a_{i, -1}&=&1&\\
@@
 	\item $a_{i,k+1-j}$: exponent of the $j$-th-to-last quantifier on line $i$. 
-	We use
-	$a_{i,-j}$ as a shorthand notation for $a_{i,k+1-j}$.
+	We use $a_{i,-j}$ as a shorthand notation for $a_{i,k+1-j}$. 
+	By indexing $a$ relative to the last quantifier instead of the first allows us to simplify
+	the linear program: for example, 
+	we do not need to keep track of the number of quantifiers on the current line. 
 	\item $x_i$: if line $i$ is the result of a speedup, it is a speedup with parameter $x_i$,
@@
 \end{enumerate}
-The constains of the type ``$su_i = b$ implies $C$'' can be encoded using the so-called ``Big M method''\footnote{See \url{https://en.wikipedia.org/wiki/Big\_M\_method}.}. 
-The constaint $d_k < d_0$ can be approximately encoded as $d_k \le d_0 - \delta$ for some small $\delta > 0$.
+The constraints of the type ``$su_i = b$ implies $C$'' can be encoded using the so-called ``Big M method''\footnote{See \url{https://en.wikipedia.org/wiki/Big\_M\_method}.}. 
+The constraint $d_k < d_0$ can be approximately encoded as $d_k \le d_0 - \delta$ for some small $\delta > 0$.
 
@@
 Using binary search on $c$, we can quickly find an upper and lower bounds on $c_k^*$.
+
+% TODO : Clean code and put online !
+An implementation of this model (and the search) are available online at \url{https://github.com/GBathie/sat-nand-lbounds}.
 

### 2022-06-27 with_mip/main.py

@@ -54,6 +54,7 @@ def run_proof(annotation, params, p0, c):
         else:
             print('Incorrect annotation')
             exit(0)
+    # Sanity check
     print('Correct proof!' if r < p0 else 'Incorrect proof :(')
 
 

### 2022-06-28 article/main.tex
Update Title
Cleaning up

@@
 %%%%%%%%%%%%%%%%%%%%%%
-\title{Uniform formula lower bounds for testing the existence of and counting satisfiable assignments}
+\title{Uniform formula lower bounds for the satisfiability problem.}
 \author{Gabriel Bathie, Ryan Williams}
@@
 \maketitle
-
-\begin{abstract}
-\end{abstract}
 
@@
 A particular attention was dedicated to the \SAT{} problem, 
-since a super-polynomial lower bound for this problem would imply $\NP \neq \P$
+since a super-polynomial lower bound for this problem would imply $\NP \neq \Poly$
 (for programs, $\NP \neq \textsf{P/poly}$ for circuits).
